Structural basis for preservation of a subset of Topologically Associating Domains in Interphase Chromosomes upon cohesin depletion

Version 1

Abstract

Compartment formation in interphase chromosomes is a result of spatial segregation between eu- and heterochromatin on a few mega base pairs (Mbp) scale. On the sub-Mbp scales, Topologically Associating Domains (TADs) appear as interacting domains along the diagonal in the Hi-C contact map (CM). Hi-C experiments showed that most of the TADs vanish upon deleting cohesin, while the compartment structure is maintained and is even enhanced. However, closer inspection of the data reveals that a non-negligible fraction of TADs is preserved (P-TADs) after cohesin loss. Imaging experiments show that, at the single-cell level, TAD-like structures are present even without cohesin. To provide a structural basis for these findings, we used polymer simulations to show that certain TADs with epigenetic mismatches across their boundaries survive after depletion of loops. More importantly, the three-dimensional structures show that many of the P-TADs have sharp physical boundaries. Informed by the simulations, we analyzed the Hi-C maps (with and without cohesin) in mouse liver and HCT-116, which affirmed that epigenetic mismatches and physical boundaries (calculated using the 3D structures) explain the origin of the P-TADs. Single-cell structures, calculated from using only the Hi-C map without any parameters, display TAD-like features in the absence of cohesin that are remarkably similar to the findings in imaging experiments, thus providing a cross validation of the computations. Some P-TADs, with physical boundaries, are relevant to the retention of enhancer-promoter/promoter-promoter interactions. Overall, our study shows that preservation of a subset of TADs upon removing cohesin is a robust phenomenon that is valid across multiple cell lines.

Introduction

Advances in experimental techniques have provided details of the three-dimensional (3D) organization of chromosomes in diverse species[l–6]. The contact map (CM) [4, 6], inferred using chromosome conformation capture technique and related variants (referred to as Hi-C from now on), is a two-dimensional (2D) matrix, whose elements indicate the probability that two loci separated by a certain genomic distance are spatially adjacent. The Hi-C experiments on different mammalian cells suggest that there are two major length scales in the organization of interphase chromosomes. On the scale, LC ∼ (2-5) Mbp (mega base pairs or Mbp), one observes checkerboard patterns in the CMs [4], which are thought to be associated with micro-phase separation between the two major epigenetic states, active (A) or euchromatin and inactive (B) or heterochromatin. On the length scale, LTAD, from tens of kb up to a few Mb, chromatin domains, referred to as Topologically Associating Domains (TADs), appear as squares along the diagonal of the CMs [7, 8]. Contacts are enriched within the TADs, and are suppressed across the boundaries between the TADs. A number of polymer models [9–14] have shown that compartment formations and TADs may be explained using micro-phase separation. The use of the two length scales, LC and LTAD, in characterizing the organization of interphase chromosomes is now entrenched in the held, although there are suggestions that finer sub-TAD structures emerge at kilo base scales [15, 16]. In particular, recent Micro-C experiments have shown that there are fine structures starting from the nucleosome level [17–19], thus establishing the hierarchical organization of interphase chromosomes over a broad range of length scales.

TADs are thought to regulate gene expression by constraining the contacts between target gene and regulatory regions [20, 21]. As a consequence, perturbation or disruption of their integrity such as deletions, duplications or inversions of DNA segments within the TADs, could lead to aberrant gene expression [8, 22–28]. A class of chromatin loops, mediated by the ATP-dependent motor cohesin [29] and the DNA-binding protein CTCF protein (“cohesin-associated CTCF loop”), organizes a subset of the TADs [30]. Cohesin [29] extrudes DNA loops of varying length, which are terminated when cohesin encounters the transcriptional insulator CCCTC-binding factor (CTCF) [31]. This implies that cohesin and CTCF are often colocalized at the TAD boundary [4–6, 30, 32, 33].

Several experiments have shown that depletion of architectural proteins (Nipbl, RAD21, and CTCF) disrupts the organization of interphase chromosomes [26, 34–41]. Schwarzer et al. [34] showed that the removal of the cohesin loading factor, Nipbl in the mouse liver cell results in loss of TADs. They concluded that compartment formation, which is independent of cohesin, is a consequence of the underlying epigenetic landscape, while TAD formation requires cohesin. Similarly, it was found that upon removal of cohesin subunit, RAD21 cohesin-associated CTCF loops and TADs are abolished [26, 39, 40]. Deletion of RAD21 results in the complete loss of the so-called loop domains [26], which are formed when CTCF colocalizes with cohesin. In contrast, imaging experiments[40] showed that TAD-like structures, with sharp boundaries are present, at the single-cell level even after deleting cohesin. Three headlines emerge from these studies are: (i) They reinforce the two-length scale description of genome organization at the ensemble level. (ii) Factors that prevent the association of cohesin with chromosomes globally abolish the TADs and the Hi-C peaks, but preserve (or even enhance) compartmentalization. Experimental studies [4, 26, 34, 39, 40] and polymer simulations [12, 35, 42] have shown that the global epigenetic state determines compartment formation, while the more dynamic TADs, with finite lifetimes[43], require ATP-dependent cohesin motor. (iii) TAD-like features persist in single cells before and after auxin treatment, albeit with changes in the locations of sharp domain boundaries.

The results of super resolution experiments [40], at the single-cell level (described above), forced us to ask if there is evidence for preservation of TADs at the ensemble level. To this end, we analyzed the experimental CMs from mouse liver and HCT-Π6 cells, in the presence and absence of cohesin to assess if TADs are preserved at the ensemble level. We discovered that, on an average a fraction of TADs, identified using the TopDom method [44], are retained in the mouse liver and HCT116 cell lines (Fig.1) after removing cohesin. These findings raise the following questions. What is the structural basis for the retention of a small but not insignificant fraction of TADs that are preserved after cohesin loss? Is it possible to establish a structural explanation for TAD retention at the ensemble level (Hi-C), which would reconcile with the results in super resolution imaging experiments showing TAD-like structures at the single level, even without cohesin?

Fate of the TADs in chromosomes upon cohesin deletion.

(a) The number of TADs in all the chromosomes, identified by the TopDom method [44], in the wild type (WT) cells and the number of preserved TADs (P-TADs) after deleting cohesin loading factor (ΔNipbl) in mouse liver. The percentage of P-TADs is ≈ 16%. (b) Same as (a) except the experimental data are analyzed for HCT-116 cell before (WT) and after RAD21 deletion. The percentage of P-TADs is 26%. (c) The total number of TADs and the number of P-TADs for each chromosome calculated using the mouse liver Hi-C data. The number above each bar is the percentage of P-TADs in each chromosome. (d) Same as (c) except the results are for chromosomes from the HCT-116 cell line. The percentage of P-TADs is greater in the HCT-116 cell line than in mouse liver for almost all the chromosomes, a feature that is more prominent in the distribution of P-TADs proportions (Right).

We answered the questions posed above by using the following strategy. We first performed polymer simulations for two chromosomes from the GM 12878 cell line using the Chromosome Copolymer Model (CCM) [11] with and without loop anchors, which roughly mimics the wild type (WT) and the absence of cohesin-associated CTCF loops, respectively. Informed by the results from the polymer simulations, we analyzed the experimental data from two cell lines [26, 34]. We discovered that epigenetic mismatch does account for a reasonable fraction (≈ 0.4 in mouse liver and ≈ 0.3 in HCT-116 cell lines) of P-TADs. We also generated the 3D structural ensemble of the chromosomes that are needed to locate the physical boundaries. The 3D structures were calculated using the parameter free HIPPS method [45] for all the chromosomes for both the cell lines. The analyses using the 3D coordinates accounted for about 53% of the P-TADs. More importantly, the 3D structures revealed TAD-like structures, at the single-cell level, exhibit in the presence and absence of cohesin, which is in accord with the super resolution imaging data [40]. Our work shows that the effects of cohesin removal on chromatin structures are nuanced, requiring analyses of both the epigenetic landscape as well as 3D structures in order to obtain a comprehensive picture of how distinct factors determine interphase chromosome organization in the nucleus. Our calculations for chromosomes from three cell lines lead to the robust conclusion that a subset of P-TADs is intact after deleting cohesin.

Results

A non-negligible fraction of TADs are preserved upon removal of cohesin

Experiments [26, 34-37, 39, 40] have shown that deletion of cohesin loaders (Nipbl in mouse liver, SCC2 in yeast), cohesin subunit (RAD21) abolishes a substantial fraction of both cohesin-associated CTCF loops and the TADs. These observations across different cell lines raise an important question: Do all the TADs completely lose their contact patterns after removal of cohesin? To answer this question, we first analyzed 5Okb-resolution CMs from the two cell lines (mouse liver [34] and HCT-116 [26]) before and after degradation of Nipbl and RAD21, respectively (see Appendix Analyses of the experimental data for details). Using TopDom [44] to identify the TADs in the CMs, we discovered that roughly 16% (26%) of TADs in mouse liver (HCT116) cell are preserved (see Fig.2 to identify P-TADs using Hi-C data) upon Nipbl (RAD21) removal (Fig.l). Approximately, 659 TADs out of 4176 are preserved (Fig.1a) after removing Nipbl in the mouse liver cells. In the HCT-116 cells, 1226 TADs out of 4733 are preserved (Fig.1b) upon RAD21 loss. Fig.1c and Fig.1d show that the number of P-TADs depends on the chromosome number. We ought to emphasize that the actual number of P-TADs, which depends on the TAD-calling protocol, is not as important as the finding that a non-negligible fraction is preserved after cohesin depletion.

Identification of preserved TADs (P-TADs) from the contact map using TopDom method.

(a) Schematic representation used to determine the P-TADs. Yellow (Blue) triangles represent the TADs identified using TopDom method in WT (cohesin depleted) contact maps at 5Okb resolution. Small square within each triangle represents a single locus (5Okb). The boundaries of a TAD detected in WT CM within ± one bin (5Okb) from a position of boundaries in cohesin depleted CM is deemed to be a P-TAD. (b) P-TAD upon cohesin loss in HCTΠ6 cell. The bar plots above the CMs show the epigenetic states. Red (Blue) color represents the active (inactive) state, respectively. The TAD between grey dashed lines is preserved upon cohesin loss. The parameter (with red square) displayed at each left bottom indicates the color scale when plotting contact maps used in Juicebox [54].

CCM simulations reproduce wild-type Hi-C maps

To explore the mechanism resulting in P-TADs, we first simulated the chromosome copolymer model [11], CCM (Appendix Fig.2a). To independently decipher the origins of P-TADs (Fig.1) in experiments, we calculated the CMs for Chr13, shown in Appendix Fig.2 (Chr10 in Appendix Fig.3) from the GM12878 cell line. The CCM simulations (Appendix Fig.2b) reproduce the ubiquitous checker board patterns well. The rectangle in Appendix Fig.2b represents the border of one such compartment formed primarily by interactions between the B-type loci.

CCM simulations reveal characteristics of P-TADs.

(a) The number of TADs in the simulated Chr10 and Chr13 chromosomes identified by the TopDom method [44] for Pl=1. The number of P-TADs after CTCF loop depletion (Pl=0) is also shown, (b) The number of P-TAD with epigenetic mismatches (blue) and those identified by the peaks in the boundary probability (green). (c)-(e) Comparison between CMs for the region of Chr13 with upper (lower) triangle with Pl = 1 (Pl = 0). The black circles at the corner of the TADs are the CTCF loop anchors. The bar above the contact map are the epigenetic states with red (blue) representing A (B) loci. Arrows above the bar show the epigenetic switch. (c) After loop deletion, TAD structures disappear. (d) TAD whose boundaries are marked by epigenetic mismatches are preserved. (e) TAD lacking at least one epigenetic mismatch is disrupted after CTCF loop loss. (f)-(h) Comparison of the CM and mean spatial-distance matrices for the 2.5Mb genomic regions (25.7-28.2Mbp,73.3-75.8Mbp and 102-104.5Mbp, respectively) with (upper) and without (lower) CTCF loop. Bottom graph shows the boundary probability, with the high values indicating population averaged TAD boundary. Purple circles in the boundary probability graph represent the preferred boundaries. A subset of P-TADs boundaries match epigenetic mismatches (blue lines). P-TADs with high boundary probability is in green line. The magenta line describes P-TADs which is not accounted by epigenetic mismatch or physical boundary in 3D space but are found using the TopDom method.

In order to quantitatively compare the Hi-C data and the simulated CMs, we transformed the contact maps to Pearson correlation maps, which is used to assess if two loci have correlated interaction profiles (Appendix Fig.2c). The Kullback-Leibler (KL) divergence between the two probability distributions for the Pearson correlation coefficients, ρĳs, from simulations and experiments is 0.04 (see Appendix Fig.2d). We also performed Principal Component Analysis (PGA) on the Pearson correlation matrix to identify the compartment structure. Comparison of the PCA-derived first principal components (PC1) across the Chr13 reveals that A/B compartments observed in the CCM correspond well to those found in the experiments (Appendix Fig.2e).

We then compared the three-dimensional spatial organization between the simulations and experiments using the Ward Linkage Matrix (WLM), which is based on an agglomerative clustering algorithm. The simulated WLM is calculated from a spatial distance map of the organized chromosome, as described in Appendix Fig.2f-h. We constructed the experimental WLM by converting the Hi-C contact map to a distance map using the approximate relationship [2, 11, 45], . Here, Pĳ is the contact probability and Rĳ is the 3D spatial distance between loci i and j (see Appendix Data Analyses). The Pearson correlation coefficient (ρwlmexp,wlmccm) between experimental and simulated WLMs is 0.83 (Appendix Fig.2h).

Snapshots of TAD structures in Appendix Fig.2k and n show that they are compact but structurally diverse. The average sizes of the TADs detected using TopDom [44] from Hi-C and simulated contact maps are ∼6l5kbs and ∼565kbs, respectively. Overall the emerging picture of the compartment and TAD structures using different methods are consistent with each other. The results in Appendix Fig.2 show that the agreement between the CCM simulations and Hi-C data is good, especially considering that (i) error estimates in the Hi-C experiments are essentially unknown, and (ii) more importantly, we did tune the value of any parameter to fit the experimental CMs. We surmise that the CCM faithfully reproduces the important features of the Hi-C maps for Chr13 (see Appendix Fig.3 for Chr10 results).

Epigenetic mismatch accounts for a large fraction of P-TADs

Most of the TADs are not discernible after loop loss, as evidenced by the blurred edges in the contact maps (Fig.3c). In the CCM simulations of chromosomes from human GM12878 cell line a subset of TADs remains even after cohesin-associated CTCF loop loss. Fig.3a shows that ≈ 44% (Chr10) and 28% (Chr13) of initial TADs identified in the Pl = 1 simulations using TopDom [44] are preserved after loop loss (Pl = 0). These values are in line with TAD preservation in chromosomes in HCT-116 [26] cells (see Fig.1d). The percentages of P-TADs depend on both the resolution of the Hi-C experiments as well as the algorithm used to identify the TADs. By using the same method to analyze both the simulation results and experimental data, it is hoped that the conclusions would be qualitatively robust.

We used the simulation results to determine the reasons for P-TADs by comparing the results for PL=l and Pl=0. The first observation is that some TADs, even with cohesin-associated CTCF loops, consist mostly of sequences in the same epigenetic state (A or B). Fig.3d compares the fate of one TAD in the region (19.3-21.8 Mb) in Chr13 between PL=l and Pl=0· The highlighted TAD is preserved upon loop loss although the probabilities of contact within this TAD is reduced when Pl = 0 (bottom) compared to Pl = 1 (top). Their boundaries correspond to a switch in the epigenetic state (the sequence location where the change in the epigenetic states occurs, as shown by the two black arrows). In contrast, a TAD in Fig.3e, that is present in the WT, is abolished when Pl=0· The disruption of this particular TAD, lacking in at least one epigenetic mismatch, occurs because it can interact more frequently with neighboring TADs composed of similar epigenetic states, which in this case is B-type loci. The importance of epigenetic mismatch in the P-TADs has been noted before [26].

The results in Figs.3d and e show that a mismatch in the epigenetic state across a TAD boundary in the WT is likely to result in its preservation after cohesin-associated CTCF loop loss. To test this assertion, we calculated the number of P-TADs that are associated with mismatches in the epigenetic states (see Appendix Data Analyses and Appendix Fig.1 for details). We considered the TAD boundary and epigenetic mismatch as overlapping if they are less than 100kb apart, which is reasonable given that Hi-C resolution adopted here is 5Okb. By using 100kb as the cut off, we only consider mismatch occurrences that exceed two loci. With this criterion, out of 216 (169) TADs calculated using TopDom 96 (47) are P-TADs for Chr13 (Chr10) (vertical blue bars in Fig.3b in which there are epigenetic mismatches in the WT).

The P-TADs with epigenetic mismatches, illustrated in Fig.3f, shows TADs in the 2.5Mbs region in Chr13. Among the three P-TADs, two of them, whose boundaries are marked by dashed blue lines, have epigenetic mismatch across the TAD boundary. These two TADs survive after removal of cohesin-associated CTCF loop.

P-TADs have prominent spatial domain boundaries

Because there are a number of P-TADs that are preserved even without epigenetic mismatches across their boundaries, we wondered if the distance matrix, which requires 3D structures, would offer additional insights about P-TADs upon cohesin-associated CTCF loop loss. Recent imaging experiments[4O, 42, 46] revealed that TAD-like domain structures with spatially segregated heterogeneous conformations are present in single cells even without cohesin. The boundaries of TAD-like domains, identified from individual 3D structures, vary from cell to cell. They exhibit a preference to reside at specific genomic positions only upon averaging, as found in the Hi-C experiments. The boundary probability at each locus is the proportion of chromosome structures in which the locus is identified as a domain boundary in the 3D space. The locations of prominent peaks in the boundary probability frequently overlap with TADs detected by the population level Hi-C maps.

To explore the relation between P-TADs in ensemble averaged CMs and preferential boundaries in individual 3D structures of chromosomes, we first constructed individual spatial distance matrices (DMs) using 10,000 simulated 3D structures, which were used to identify the single-cell domain boundaries (see Appendix Data Analyses for details and Appendix Fig.6). We find preferential domains, with high peaks in the boundary probability along the genomic region as well as variations in single-cell domains both in PL = 1 and Pl = 0 (see Appendix Fig.8).

The simulations show that TADs with epigenetic mismatches across the boundary are likely to be preserved after cohesin-associated CTCF loop loss. Furthermore, Fig.3f (blue dashed lines) shows that single-cell domain boundaries preferentially reside at the TAD boundaries with epigenetic mismatches, leading to a prominent signature for the ensemble boundary after averaging over a population of cells. Interestingly, the P-TAD has prominent peaks in boundary probability, even without epigenetic mismatch, at the same genomic position as in the CM (green lines in Fig.3f and g). These observations imply that the presence of physical boundaries in the 3D structures may be used to identify P-TADs, especially in instances when there are no epigenetic mismatches.

Not all preferential boundaries identified in the distance matrices of the WT cells coincide with the TADs detected using the CM [40]. There is discordance in the TAD boundaries and high peaks in boundary probability. The top panel in Fig.3h (magenta lines) shows that in the (102 - 104.5) Mb range, TopDom predicts that there are three P-TADs in the WT after loop loss (see the top panel with Pl = 0). There are two prominent peaks in the WT boundary probability whose boundaries coincide with the TADs predicted by TopDom (see bottom panel with Pl=1 in Fig.3h). But the peak height for the third TAD is very small. At best, one can deduce from the boundary probabilities (compare the results in Fig.3h for Pl=1 and Pl=0) that the middle TAD is preserved, which would be consistent with the TopDom prediction.

We calculated a standardized Z-score for the boundary probability in the genomic region to determine the preferred boundaries in single-cell domains. The number of P-TADs that are accounted for by prominent boundary peaks increases if Z-score is reduced. This implies that some P-TADs detected in the CMs using TopDom have weak physical boundaries in the 3D structures. We considered the maxima, with Z-score values larger than 0.7, as preferred boundaries in order to determine if P-TADs arise due to the presence of strong physical boundary. With this criterion, we obtained good agreement for the mean length of the TADs detected in the CM using the TopDom method. The averaged sizes of the TADs in Chr13 using TopDom and boundary probability are ∼565kbs and ∼535kbs, respectively. Quantitative analysis of the boundary probabilities along the genomic region revealed ≈66% of the P-TADs in Chr10 and Chr13 have preferential positioning in single-cell domains (green bars in Fig. 3b). Most P-TADs with epigenetic mismatches display prominent peaks in the boundary probability (≈ 85%).

Let us summarize the primary lessons from the simulations, which form the basis for analyzing the experiments on chromosomes from mouse liver and HCT-116 cell lines. (i) Mismatch in the epigenetic state across the TAD boundary is a predominant factor in determining the P-TADs after CTCF loop deletion. (ii) The presence of physical boundaries in the 3D structures accounts for certain fraction of P-TADs, although in some instances TopDom predictions (used in (i)) are not compatible with boundaries deduced from 3D structures.

Structural explanation of P-TADs upon cohesin removal from analysis of Hi-C data

In order to assess if the conclusions from simulations explain the experimental data, we first calculated the number of P-TADs whose boundaries have switches in the A/B epigenetic states in mouse liver and HCT-116 cell lines. We assigned chromatin state A (active, red) or B (repressive, blue) by analyzing the combinatorial patterns of histone marks using ChromHMM [47] (see Appendix analyses of the experimental data and Appendix Fig.7). An average over the 20 chromosomes shows that 280 P-TADs were associated with a mismatch between A and B epigenetic states upon ΔNipbl in mouse liver (blue bar in Fig.4a). The corresponding number of P-TADs, averaged over 23 chromosomes, with epigenetic mismatches is 396 after deleting RAD21 in HCT-116 (blue bar in Fig.4b). Not unexpectedly, TADs with epigenetic mismatches across their boundaries are preserved with high probability. We find that a large number of P-TADs are accounted (green bars in Fig.4a and Fig.4b) for by the presence of peaks in the boundary probabilities, which we discuss in detail below.

We then searched for a structural explanation for P-TADs in the two cell lines (Fig.4a and Fig.4b). A plausible hint comes from the the CCM simulations (Fig. 3), which show that boundary probabilities are good predictors of P-TADs. This implies that peaks in the boundary probabilities should correspond to P-TADs. Similar findings are obtained by analyzing the experimental data (Fig. 4c and d). Additional examples are discussed in Fig. 5. In light of these findings, we wondered if, in general, physical boundaries can be inferred directly using data from ensemble experiments. To this end, we used the HIPPS method (see Methods), to calculate an ensemble of 3D structures with the Hi-C contact map as the input. Several conclusions follow from the results in Fig.5. First, Figs. 5a and b show that HIPPS faithfully reproduces Hi-C contact maps. Second, using the ensemble of 3D structures, we calculated the locus-dependent boundary probabilities for both the WT and cohesin-depleted cells. Comparison of the peak positions in the averaged boundary probabilities and TAD boundaries shows that they often coincide, although there are discordances as well (Fig.5 and Appendix Figs. 12 and 13). Third, when there is a mismatch in the epigenetic states, a substantial fraction of P-TADs have high peaks in the boundary probabilities (see Fig.5 and Appendix Fig. 12). As in the simulations, a large fraction of P-TADs (≈ 67%) have high peaks in the boundary probabilities. Taken together, the results show that the predictions using boundary probabilities and the TopDom method are consistent. Finally, analyses of the experiments suggest that the epigenetic state as well as the presence of physical boundary in the 3D structures have to be combined in order to determine the origin of P-TADs in cohesin depleted cells.

Classification of P-TADs from Hi-C maps from two cell lines and link between boundary probability peak and epigenetic mismatch.

(a) The number of P-TADs in all the chromosomes (orange bar taken from Fig. la) that are accounted for by epigenetic mismatches (blue bar) as well as peaks in the boundary probability (green bar) after Nipbl loss in mouse liver. (b) Same as (a) except the analyses is done using experimental data are for HCT-116 cell after ΔRAD21. (c) Example of P-TAD in the WT 97.7Mb-100.2Mb region of Chr3 from HCT-116 cell line. The CM resolution is 5Okb.The mean DMs calculated using the 3D structures is in the middle panel. The dark-red circles at the boundaries of the TADs in the CMs are loop anchors detected using HiCCUPS [55]. The peaks in the boundary probability (bottom panel) are shown by purple circles. Epigenetic mismatch coincides with peak in the boundary probability (compare top and bottom panels). Bottom plot shows the probability for each genomic position to be a single-cell domain boundary. (d) Same as (c) except the results correspond to the absence of RAD21. Although not as sharp, there is discernible peak in the boundary probability when there is an epigenetic mismatch after removal of RAD21.

Fate of TADs after ΔNipbl in mouse liver cells.

(a)-(b) Comparison between Hi-C CMs (lower) and calculated CMs (upper) using the 3D structures obtained from HIPPS method for the 3Mb genomic regions (Chr6:22.6Mb-26.IMb in WT cells and Chr7:l39Mb-l42.5Mb in Nipbl-depleted cells), respectively. The distance threshold for calculated CMs is adjusted to achieve the best agreement between HIPPS and experiments. Calculated CMs are in very good agreement with Hi-C CMs in both WT and Nipbl-depleted cells. (c) Complete Loss (Chr6:23.55Mb-26.O5Mb) of TADs. (d)-(e) P-TADs (Chr7:l39.5Mb-l42Mb and Chr15:89.5Mb-92Mb). In all cases, the plots below the scale on top, identifying the epigenetic states[47], compare 5Okb-resolution Hi-C CMs for the genomic regions of interest with Nipbl (upper) and without Nipbl (lower). Mean spatial-distance matrices, obtained from the Hi-C contact matrices using the HIPPS method[45], are below the CMs. The dark-red circles at the boundaries of the TADs in the CMs are loop anchors detected using HiCCUPS [4]. ChIP-seq tracks for CTCF, RAD21 and SMC3 in the WT cells [34] illustrate the correspondence between the locations of the most detected loop anchors and the ChIP-seq signals. Bottom plots give the probabilities that each genomic position is at a single-cell domain boundary in the specified regions. Purple circles in the boundary probability graph represent the physical boundaries. A subset of physical boundaries in P-TADs coincide with epigenetic mismatches (blue lines), indicating that the probabilities of contact at these boundaries are small. P-TADs in (e), demarcated by green lines, have high peaks in the boundary probability without epigenetic mismatch.

With the near quantitative agreement with experiments, we performed detailed analyses, based on the epigenetic mismatches and boundary probabilities for chromosomes from the mouse liver (Fig.5). The Appendix contains analyses of the experimental data, and the results for HCT-116 cell are given in Appendix Fig. 12. To illustrate different scenarios, we consider the 2.5Mbs regions from Chr6 (Fig.5c), Chr7 (Fig.5d), and Chr15 (Fig.5e). (i) For Chr 6, there are three TADs according to TopDom (Fig.5c) in the WT. Upon ΔNipbl, these TADs are abolished (compare the top and bottom panels in Fig 5c). The epigenetic track indicates that the region is mostly in the repressive (B) state. Quantification of the boundary probabilities along the 2.5Mb region of Chr6 shows that the TADs also lack physical boundaries upon ŠNipbl. (ii) Examples of P-TADs that satisfy the epigenetic mismatch criterion are given in Fig.5d. Using TopDom, we identified several TADs (top panel in Fig.5d) in this region of Chr7. It is interesting that the boundary probabilities obtained from the HIPPS-generated distance matrices are also large when there is a mismatch in the epigenetic states. In these examples, both epigenetic mismatches and boundary probabilities give consistent results (see the dashed blue lines in Fig.5d). Two TADs in the WT (the ones on the right in the upper panel in Fig.5d) merge to form a single TAD in the ΔNipbl. This observation is in accord with expectation based on epigenetic mismatch, whose corollary is that if there is a TAD within a region that contains predominantly A or B type loci they ought to merge upon ΔNipbl. (iii) In the 2.5Mb region of Chr15, there are three TADs in the WT (top panel in Fig.5e). The first and the third TADs have an epigenetic mismatch at only one boundary (blue dashed line), and the expectation is that they would not be preserved upon Nipbl removal. However, the boundary probabilities show that the TADs have physical boundaries in both, and thus they are preserved. Taken together, the results in Fig.5 show that only by combining the epigenetic mismatches (Hi-C data is sufficient) and the boundary probabilities (3D structures are required), one can account for a number of P-TADs.

Single-cell structural change upon cohesin depletion

Finally, we asked whether HIPPS can capture the 3D structural changes in cohesin-depleted cell at the single-cell level. To this end, we compare the structures obtained using HIPPS with the imaging data [40], which examined the consequences of ΔRAD21 in HCT-Π6. We used HIPPS to the Hi-C contact map on the same genomic region as in the experiment [40] to generate the 3D structures. The results of our calculations for a 2.5Mbp in Chr21 (34.6-37.1Mb) region from HCT-Π6 cell line for the WT and ΔRAD21 are presented in Fig.6. The DMs were calculated from the calculated 3D structures using the HIPPS method (Fig.6a). The mean DMs for the WT and ΔRAD21 are shown on the left and right panels in Fig.6b. Similar results for Chr 4 in mouse liver cell (Chr2 in HCT-116) are displayed in Appendix Fig.9 (Appendix Fig. 10). Several inferences may be drawn from the results in Fig.6. (1) There are large variations in the distance matrices and domain boundary locations/strengths from cell to cell (Fig.6c and d). This finding is in excellent agreement with imaging data [40]. (2) In both experiments and our calculations, there are TAD-like structures at the single level even after RAD2I is removed (see the right panel in Fig.6c for the theoretical predictions). We should add that TAD-like structures in single cells with and without cohesin have also been found using the strings and binders polymer model [12]. (3) The calculated boundary strength distribution (blue histogram in the left panel in Fig.6d) for the WT is in reasonable agreement with the measured distribution (purple histogram from [40]). Similarly, the calculated and measured boundary strength distributions for ΔRAD21 cells are also in good agreement (right panel in Fig.6d). Just as in experiments [40], we find that the distributions of boundary strength are the same in the WT and in cells without RAD21. (4) We also find that the average locus-dependent boundary probability obtained using theory is in very good agreement with the reported experimental data (compare the curves in the left panel in Fig.6e for the WT and the ones on the right for ΔRAD21 cells).

Calculated 3D structures from Hi-C produce TAD-like structures found in imaging experiments

On the left (right) panels are results for WT (ΔRAD21) for (Chr21:34.6-37.1Mb). For visualization purposes we adopted the color scheme used in the imaging study [40]. (a) Hi-C contact maps with (left) and without ΔRAD21 [26]. (b) Mean distance matrices calculated from HIPPS-generated 3D strictures. (c) Examples of calculated single-cell distance matrices with (left) and without (right) RAD21. Schematic of structures for the two cells under the two conditions are given below. (d) Distribution of the boundary strengths, describing the steepness in the changes in the spatial distance across the boundaries. The left (right) is for the WT (ΔRAD21) cells. The blue (purple) histogram is calculated using HIPPS (experiments). (e) Position dependent boundary probability for the WT (left) RAD2I deleted cells (right). The curve in blue (purple) is the calculated (measured) boundary probability for the WT cells. The orange (green) curve is from the calculations (experiments). The plots show that the location of prominent peaks in the calculated boundary probability are in excellent agreement with experiments for the WT cells (left panel). Without RAD2I, high peaks are absent in both the cases (right panel). The calculated results were obtained without adjusting any parameter in the theory.

Discussion

By analyzing the experimental Hi-C data [26, 34], we showed that upon cohesin loss a non-negligible fraction of TADs is preserved. To examine the factors that control the P-TADs, we first performed polymer simulations of two chromosomes in the presence and absence of cohesin-associated CTCF loops from the GM 12878 cell line. The simulations for chromosomes from a cell line reproduced the salient findings noted in the experiments in mouse liver and HCT-Π6 cells [26, 34]. The key findings in the simulations are: (1) Mismatches in the epigenetic states across the TAD boundary account for a large fraction of P-TADs. Certain P-TADs, that lack epigenetic mismatches, could be preserved as revealed by the presence of physical boundaries in the 3D structures. (2) In general, there is an overall consistency between predictions based on boundary probability calculations using 3D structures, and mismatches in the epigenetic switches across TAD boundaries.

Informed by the simulation results, we analyzed the experimental Hi-C data on chromosomes from two cell lines [26, 34]. Analyses of the structures determined from the Hi-C map using the HIPPS [45], with and without Nipbl or RAD21, showed that A-type loci form larger spatial clusters after cohesin removal, consistent with enhancement of compartments inferred from Hi-C CMs (Appendix Fig.4). More importantly, we found that a large fraction of P-TADs can be accounted for by the epigenetic mismatches and the preferential boundaries in 3D structures of chromosomes. Thus, a combination of the two perspectives accounts for the experimental observations pertaining to P-TADs upon cohesin loss[26, 34]. Most of the P-TADs with epigenetic mismatches in the CMs have a prominent peak in the boundary probability, which indicates a strong physical boundary in the 3D structures. If this interpretation is correct, then a natural conclusion is that not only micro-phase separation on the larger scale, LC but also some special TADs on a shorter scale, Ltad are encoded in the epigenetic sequence.

P-TADs are due to enhancer/promoter interactions

Cohesin is thought to directly or indirectly regulate enhancer-promoter (E-P) interactions. However, strikingly a recent Micro-C experiment discovered that E-P and promoter-promoter (P-P) interactions are largely insensitive to acute depletion of cohesin [48]. It has been previously shown that E-P/P-P interactions form one or multiple self-associating domains, strips extending from domain borders and loop-like structures at their intersections at a finer scale [4951]. Based on the recent finding [48], we explored if maintenance of finer-scale E-P and P-P interactions is the origin of the P-TADs without epigenetic mismatches by analyzing Micro-C data [48]. The left panel in Fig.7a shows cohesin-associated (green dashed line) and cohesin-independent (blue dashed line) TAD structures (defined using Topdom) in the WT cells. In the latter case, the TADs E-P and P-P loops (blue circles) exist at the boundary but there is no epigenetic mismatch, implying that it is a domain needed for E-P or P-P communication. Interestingly, the TADs were also conserved upon cohesin loss (right panel in Fig.7a). Analyses of the 3D structures (Fig.7b) also reveal that the TADs with E-P/P-P loops have strong physical boundaries even in the absence of cohesin. Fig.7c shows an example of a TAD with both E-P/P-P loops and cohesin/CTCF loops at the boundary in the WT cells is retained after cohesin deletion, and is associated with prominent boundary peaks. We propose that only a subset of TADs are conserved, potentially for functional reasons. Taken together, our observations suggest that maintenance of E-P/P-P interactions is the origin of the P-TADs even in the absence of epigenetic mismatches. It is worth emphasizing that these conclusions can only be obtained by analyzing the 3D structures, which we calculated from the Micro-C contact maps using the parameter free HIPPS method [45].

Certain TADs enriched in E-P/P-P interactions at the boundary are preserved upon cohesin deletion.

(a) Comparison between 5kb Micro-C CMs in the region (Chr8:72.24Mb-72.57Mb) in the WT (left panel) and cohesin-depleted (right panel) mESC cells [48]. The location of cohesin loops (green square) and E-P/P-P (blue circles) plotted in the WT CMs are from experiments[48]. Each bar above contact map shows epigenetic states (Red: Active, Blue: Inactive) annotated based on ChromHMM results [56]. The cohesin-dependent (green dashed lines) and independent (blue dashed lines) TADs are detected in the WT cells using the TopDom method with default parameter (w=5). P-TADs (blue dashed lines) are also found in cohesin deleted cells. (b)-(c) Comparison between 2Okb Micro-C CMs and mean DMs spanning the regions (Chr19:8.66-9.2Mb and Chr12:56.4-56.9Mb, respectively.) in the presence (upper) and absence (lower) cohesin. Bottom graph shows below each DM shows the boundary probability calculated from 10,000 3D structures calculated using the parameter free HIPPS method. P-TADs between grey dashed lines are detected using TopDom method (w=5). A P-TAD with high boundary peak, without epigenetic mismatches, are enriched due to E-P/P-P interactions at the boundaries.

TAD-like structure are preserved at the single-cell level

Remarkably, the conclusion that there are cell-to-cell variations in the distance maps, noted in imaging experiments [40], are also fully affirmed in the calculated 3D structures calculated directly from the ensemble Hi-C data. In addition, the calculations also show that in mouse liver and HCT-116 cell lines, TAD-like structures are present after removing Nipbl and RAD21, respectively. We established that the Hi-C data sets can be utilized to generate 3D structures using HIPPS, which allows us to examine the impact of perturbations on three-dimensional organization of interphase chromosomes at the single-cell level. These results are significant because (1) only a limited number of loci can be directly imaged whereas Hi-C data can be routinely generated at higher resolution, (2) the number of Hi-C data on various cell types and species currently is greater than that from imaging data.

Comments on the Methods

In order to explore the factors that control the P-TADs, there are two assumptions. (1) The results of the Hi-C experiments are taken at face value in the sense. We view this as an assumption because errors in the Hi-C readouts may be difficult to evaluate even though such experiments are invaluable [52]. (2) The TADs were identified using TopDom, one of many TAD callers. A recent survey [53] shows that, although the finding that TADs are hierarchically organized domains is robust, there are substantial variations in the identification of these domains predicted by different methods. Although TopDom fairs reasonably well in comparison to other methods, there is no guarantee that it identifies the TAD location or the number of TADs accurately. It is only for convenience that we used TopDom as the reference to which the results using the boundary probabilities are compared.

eLife Assessment

This valuable study, of interest for students of the biology of genomes, uses simulations in combination with published data to examine how many TADs remain after cohesin depletion. The authors suggest that a significant subset of chromosome conformations do not require cohesin, and that knowledge of specific epigenetic states can be used to identify regions of the genome that still interact in the absence of cohesin. The theoretical approaches and quantitative analysis are state-of-the-art, and the data quality and strength of the conclusions are solid. However, because "physical boundaries (of domains?)" in the model appear to be a consequence of preserved TADs, rather than the other way around, the functional insights are limited.

Reviewer #1 (Public Review):

Summary:

In this paper, Jeong et al. investigate the prevalence and cause of TADs that are preserved in eukaryotic cells after cohesin depletion. The authors perform an extensive analysis of previously published Hi-C data, and find that roughly 15% of TADs are preserved in both mouse liver cells and in HCT-116 cells. They confirm previous findings that epigenetic mismatches across the boundaries of TADs can cause TAD preservation. However, the authors also find that not all preserved TADs can be explained this way. Jeong et al. provide an argument based on polymer simulations that "physical boundaries" in 3D structures provide an additional mechanism that can lead to TAD preservation. However, in its current form, we do not find the argumentation and evidence that leads to this claim to be fully compelling.

Strengths:

We appreciate the extensive statistical analysis performed by the authors on the extent to which TAD's are preserved; this does seem like a novel and valuable contribution to the field.

Weaknesses:

1. As the authors briefly note, the fact that compartmentalization due to epigenetic mismatches can cause TAD-like structures upon cohesin depletion has already been discussed in the literature; see for example Extended Data Figure 8 in (Schwarzer et al., 2017) or the simulation study (Nuebler et al., 2018). We are hence left with the impression that the novelty of this finding is somewhat overstated in this manuscript.

2. It is not quite clear what the authors conceptually mean by "physical boundaries" and how this could offer additional insight into preserved TADs. First, the authors use the CCM model to show that TAD boundaries correlate with peaks in the single cell boundary probability distribution of the model. This finding is consistent with previous reports that TAD-like structures are present in single cells, and that specific TAD boundaries only arise as a population average. The finding based on the CCM simulations hence seems to be that preserved TADs also arise as a population average in cohesin-depleted cells, but we do not follow what the term "physical boundaries" refers to in this context. The authors then use the Hi-C data to infer a maximum-entropy-based HIPPS model. They find that preserved TADs often have boundaries that correspond to peaks in the single cell boundary probabilities of the inferred model. The authors seem to imply that these peaks in the boundary probability correspond to "physical boundaries" that cause the preservation of TADs. This argument seems circular; the model is based on inferring interaction strengths between monomers, such that the model recreates the input Hi-C map. This means that the ensemble average of the model should have a TAD boundary where one is present in the input Hi-C data. A TAD boundary in the Hi-C data would then seem to imply a peak in the model's single-cell boundary probability. (The authors do display two examples where this is not the case in Fig.3h, but looking at these cases by eye, they do not seem to correspond to strong TAD boundaries.) "Physical boundaries" in the model are hence a consequence of the preserved TADs, rather than the other way around, as the authors seem to suggest. At the very least the boundary probability in the HIPPS model is not an independent statistic from the Hi-C map (on which their model is constrained), so we have concerns about using the physical boundaries idea to understand where some of the preserved TADs come from.

References:

Nuebler, J., Fudenberg, G., Imakaev, M., Abdennur, N., & Mirny, L. A. (2018). Chromatin organization by an interplay of loop extrusion and compartmental segregation. Proceedings of the National Academy of Sciences of the United States of America, 115(29), E6697-E6706. https://doi.org/10.1073/PNAS.1717730115/SUPPL_FILE/PNAS.1717730115.SAPP.PDF

Schwarzer, W., Abdennur, N., Goloborodko, A., Pekowska, A., Fudenberg, G., Loe-Mie, Y., Fonseca, N. A., Huber, W., Haering, C. H., Mirny, L., & Spitz, F. (2017). Two independent modes of chromatin organization revealed by cohesin removal. Nature 2017 551:7678, 551(7678), 51-56. https://doi.org/10.1038/nature24281

Reviewer #2 (Public Review):

Summary:

Here Jeong et al., use a combination of theoretical and experimental approaches to define molecular contexts that support specific chromatin conformations. They seek to define features that are associated with TADs that are retained after cohesin depletion (the authors refer to these TADs as P-TADs). They were motivated by differences between single cell data, which suggest that some TADs can be maintained in the absence of cohesin, whereas ensemble HiC data suggest complete loss of TADs. By reananalyzing a number of HiC datasets from different cell types, the authors observe that in ensemble methods, a significant subset of TADs are retained. They observe that P-TADs are associated with mismatches in epigenetic state across TAD boundaries. They further observe that "physical boundaries" are associated with P-TAD maintenance. Their structure/simulation based approach appears to be a powerful means to generate 3D structures from ensemble HiC data, and provide chromosome conformations that mimic the data from single-cell based experiments. Their results also challenge current dogma in the field about epigenetic state being more related to compartment formation rather than TAD boundaries. Their analysis is particularly important because limited amounts of imaging data are presently available for defining chromosome structure at the single-molecule level, however, vast amounts of HiC and ChIP-seq data are available. By using HiC data to generate high quality simulated structural data, they overcome this limitation. Overall, this manuscript is important for understanding chromosome organization, particularly for contacts that do not require cohesin for their maintenance, and for understanding how different levels of chromosome organization may be interconnected. I cannot comment on the validity of the provided simulation methods and hope that another reviewer is qualified to do this.

Reviewer #3 (Public Review):

This manuscript presents a comprehensive investigation into the mechanisms that explain the presence of TADs (P-TADs) in cells where cohesin has been removed. In particular, to study TADs in wildtype and cohesin depleted cells, the authors use a combination of polymer simulations to predict whole chromosome structures de novo and from Hi-C data. Interestingly, they find that those TADs that survive cohesin removal contain a switch in epigenetic marks (from compartment A to B or B to A) at the boundary. Additionally, they find that the P-TADs are needed to retain enhancer-promoter and promoter-promoter interactions.

Overall, the study is well-executed, and the evidence found provides interesting insights into genome folding and interpretations of conflicting results on the role of cohesin on TAD formation.

To strengthen their claims, the authors should compare their de-novo prediction approach to their data-driven predictions at the single cell level.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment:

This valuable study, of interest for students of the biology of genomes, uses simulations in combination with published data to examine how many TADs remain after cohesin depletion. The authors suggest that a significant subset of chromosome conformations do not require cohesin, and that knowledge of specific epigenetic states can be used to identify regions of the genome that still interact in the absence of cohesin. The theoretical approaches and quantitative analysis are state-of-the-art, and the data quality and strength of the conclusions are solid. However, because "physical boundaries (of domains?)" in the model appear to be a consequence of preserved TADs, rather than the other way around, the functional insights are limited.

Summary of the reviewer discussion for the authors:

While the simulations are state of the art and the reviewers appreciated that the approaches used here might help to resolve apparent discrepancies between conclusions from single-cell and bulk/ensemble techniques to study chromosome conformation, the work would benefit from clarification of what precisely is meant with "physical boundaries" and from a comparison of CCM and HIPPS models to understand commonalities and differences between them. In addition, more discussion of the relation of the current work to previous studies, such as Schwarzer et al., 2017, and Nuebler et al., 2018, would elevate the work and make the key claims more compelling. Please see also the detailed comments from the expert reviewers.

We thank the editor for the assessment and the reviewers for the incisive comments. We will address these comments one by one. In particular, we attempt to clarify the concept of “physical boundaries” and its relevance in our study. We hope our responses are satisfactory. We believe that our manuscript has benefitted substantially by revising the manuscript following the comments by the reviewers.

Below is our point-by-point response to the comments:

Reviewer #1 (Public Review):

Summary:

In this paper, Jeong et al. investigate the prevalence and cause of TADs that are preserved in eukaryotic cells after cohesin depletion. The authors perform an extensive analysis of previously published Hi-C data, and find that roughly 15% of TADs are preserved in both mouse liver cells and in HCT-116 cells. They confirm previous findings that epigenetic mismatches across the boundaries of TADs can cause TAD preservation. However, the authors also find that not all preserved TADs can be explained this way. Jeong et al. provide an argument based on polymer simulations that "physical boundaries" in 3D structures provide an additional mechanism that can lead to TAD preservation. However, in its current form, we do not find the argumentation and evidence that leads to this claim to be fully compelling.

Strengths:

We appreciate the extensive statistical analysis performed by the authors on the extent to which TAD's are preserved; this does seem like a novel and valuable contribution to the field.

We thank the reviewer for a succinct and comprehensive summary of our work and for appreciating value of our work.

Weaknesses:

1. As the authors briefly note, the fact that compartmentalization due to epigenetic mismatches can cause TAD-like structures upon cohesin depletion has already been discussed in the literature; see for example Extended Data Figure 8 in (Schwarzer et al., 2017) or the simulation study (Nuebler et al., 2018). We are hence left with the impression that the novelty of this finding is somewhat overstated in this manuscript.

It is unclear to us by studying the results in the Extended Data Figure 8 that the authors have shown that epigenetic mismatches cause TAD-like structures. As far as we can discern, the data, without a quantitative analysis, shows that may be new TAD-like structures that are not in the wild type appear when cohesin is deleted.

The studies by Schwarzer et al 2017 and Nuebler et al 2018 are relevant to our own investigation, which we undertook after scrutinizing the experiments in Schwarzer et al 2017 and the related work by Rao et. al in 2017 on a different cell line. In the summary of the Reviewer discussion, it is suggested we discuss the relation to the experimental study by Schwarzer et al 2017 and the computational work by Nuebler et al 2018.

(1) The results and the corresponding discussion in these two studies are different (may be complimentary) from our results. When referring to the Extended Data Figure 8 Schwarzer and co-authors state in the main text, “The finer compartmentalization explains most of the remaining or new domains and boundaries seen in Nipbl Hi-C maps”. We are not 100% sure what “remaining” means in this context. The Extended Data Fig. 8(a) shows the “common boundaries” is correlated with the eigenvectors of compartmentalization. If this indeed is what the reviewer is referring to, we believe that our study differs from theirs in two important ways: First, Extended Data Fig.8 (a) is a statistical analysis at the “ensemble” level. In our study, we examined the preservation of TADs at both individual and ensemble level with more detailed analysis. Second, in Extended Data Fig. 8(a), the “common boundaries” (incidentally we are uncertain how that was calculated) are compared to the eigenvectors of PCA analysis of the compartments (larger length scales). In contrast, in our study, we examined the correlation between TAD boundaries and the epigenetic profiles. We believe that this is an important distinction. The PCA analysis of compartments and “common boundaries” defined using (presumably) the insulation score are both derived from the Hi-C contact map. Epigenetic profile, on the other hand, is independent of Hi-C data. We believe our contribution, is to build the connection between epigenetic profiles with the preservation of TADs, and link it to 3D structures. For these reasons, we assert that our results are novel, and are not contained (or even implied) in the Schwarzer et al 2017 study.

The simulations in Neubler et al 2018, which were undertaken to rationalize the experimenrs, revealed that compartmentalization of small segments is enhanced after cohesin depletion, while TADs disappear, which support the broad claims that are made in the experiments. They assert that the structures generated are non-equilibrium. They do not address the emergence of preserved nor the observation of TAD-like structures at the single cell level. However, our goal was to elucidate the reasons for of preservation of TADs. By that we mean, the reasons why certain TADs are present in both the wild and cohesin depleted cells? Through a detailed analyses of two cells, polymer simulations, we have provided a structural basis to answer the question. Finally, we have provided a plausible between TAD preservation and maintenance of enhancer-promoter interactions by analyzing the Micro-C data. For all these reasons, we believe that our study is different from the results in the Extended Figure 8 or the simulations described by Neubler.

Let us summarize the new results in our study that are not contained in the studies referred to by this Reviewer. (1) We showed by analyzing the Hi-C data for mouse liver and HCT-16 that a non-negligible fraction of TAPs is preserved, which set in motion our detailed investigation. (2) Then, using polymer simulations on a different cell type, we generated quantitative insights (epigenetic mismatches as well as structural basis) for the preservation of TADs. Although not emphasized, we showed that deletion of cohesin in the GM12878 cells also give rise to P-TADs a prediction that suggests that the observations might be “universal”. (3) Rather than perform, time consuming polymer simulations, we calculated 3D structures directly from Hi-C data for the mouse liver and HCT-16 cells, which provided a structural basis for TAP preservation. (4) The 3D structures also showed how TAD-like features appear at the single cell level, which is in accord with imaging experiments. (5) Finally, we suggest that P-TADs may be linked to the maintenance of enhancer-promoter and promoter-promoter interactions by calculating the 3D structures using the recent Micro-C data.

For the reasons given above, we assert that our results are novel, and bring new perspectives that are not in the aforementioned insightful studies cited by the Reviewer.

1. It is not quite clear what the authors conceptually mean by "physical boundaries" and how this could offer additional insight into preserved TADs. First, the authors use the CCM model to show that TAD boundaries correlate with peaks in the single cell boundary probability distribution of the model. This finding is consistent with previous reports that TAD-like structures are present in single cells, and that specific TAD boundaries only arise as a population average.

The finding based on the CCM simulations hence seems to be that preserved TADs also arise as a population average in cohesin-depleted cells, but we do not follow what the term "physical boundaries" refers to in this context. The authors then use the Hi-C data to infer a maximumentropy-based HIPPS model. They find that preserved TADs often have boundaries that correspond to peaks in the single cell boundary probabilities of the inferred model. The authors seem to imply that these peaks in the boundary probability correspond to "physical boundaries" that cause the preservation of TADs. This argument seems circular; the model is based on inferring interaction strengths between monomers, such that the model recreates the input Hi-C map. This means that the ensemble average of the model should have a TAD boundary where one is present in the input Hi-C data. A TAD boundary in the Hi-C data would then seem to imply a peak in the model's single-cell boundary probability. (The authors do display two examples where this is not the case in Fig.3h, but looking at these cases by eye, they do not seem to correspond to strong TAD boundaries.) "Physical boundaries" in the model are hence a consequence of the preserved TADs, rather than the other way around, as the authors seem to suggest. At the very least the boundary probability in the HIPPS model is not an independent statistic from the Hi-C map (on which their model is constrained), so we have concerns about using the physical boundaries idea to understand where some of the preserved TADs come from.

There are many statements in this long comment that require us to unpack separately. First, using both the CCM simulations, and even more importantly using data-driven approach, we established that TAD-like structures are present in single cells with and without cohesin. The latter finding is fully consistent with imaging experiments. We are unaware of other computational efforts, before our work, demonstrating that this is the case. Perhaps, the Reviewer can point to the papers in the literature.

Regarding the statement that our arguments are circular, and lack of clarity of the meaning of physical boundary, please allow us to explain. First, we apologize for the confusion. Let us clarify our approach. We first used CCM to understand the potential origin of substantial fraction of P-TADs in the GM. The simulations, allowed us to generate the plausible mechanisms, for the origin of P-TADs. Because the CCM does reproduce the Hi-C data, we surmised that the general mechanisms inferred from these simulations could be profitably used to analyze the experiments. The simulations also showed that knowledge of 3D structures produces a muchneeded structural basis of P-TADs , and potentially for emergence of new TADs upon cohesin depletion.

Because 3D coordinates are needed to obtain structural insights into the role of cohesin, we use a novel method to obtain them without the need for simulations. In particular, we used the HIPPS method to obtain 3D coordinates with the Hi-C data as the sole input, which allowed us to calculate directly the boundary probabilities. The excellent agreement between the predicted 3D structures and imaging experiments suggests that meaningful information, not available in Hi-C, may be gleaned from the ensemble of calculated 3D structures.

Although “physical boundary”, a notion introduced by Zhuang, is defined in in the method section, it is apparently unclear for which we apologize. Because this is an important technical tool, we have added a summary in the main text in the revision. We did not mean to imply that the physical boundaries cause the preservation of TADs, although we found that maintenance of the enhancer-promoter contacts (see Fig. 8 in the revision) could be one of the potential reasons for the emergence of physical boundaries. We agree with the reviewer that physical boundaries are structural evidence of preserved TADs (not the cause), that is when a TAD is preserved, we can detect it by prominent physical boundary. The purpose and benefit of physical boundary analysis and using HIPPS in general is to obtain three-dimensional structures of chromosomes. Although both CCM simulations and HIPPS use Hi-C contact maps, three-dimensional structures provide additional information that is not present in the Hi-C data.

The arguments that the authors use to justify their claims could be clarified and strengthened. Here are some suggestions: -Explain the concept of "physical boundaries" more clearly in the main text.

As explained above, we have revised the text to clarify the concept and purpose of physical boundaries analysis. See Page 7.

Justify why the boundary probabilities and the physical boundaries concept can be used to offer novel insight into where preserved TADs may come from.

Boundary probabilities and physical boundaries provide previously unavailable 3D structural information on the TADs structures both at the single-cell and population level. This provides a direct structural basis for determining which TADs are preserved. But in order to understand where P-TADs may come from, physical boundaries analysis alone is not sufficient. As we have shown in the analysis of enhancer-promoter contact, using physical boundary analysis from 3D structures, we can conclude that conservation of enhancer-promoter contact could be one of the reasons for the P-TAD.

Explain more clearly what the additional value of using the HIPPS model to study TAD preservation is.

Our goal, as announced in the title is to elucidate the structural basis for the emergence of PTADs. The HIPPS method, which avoids doing simulations (like CCM and other polymer models used in the literature) provides an ensemble of 3D conformations using averaged contact map generated in Hi-C experiments. Even more importantly, HIPPS produce an ensemble of structures, which can be the basis for predicting the outcomes at the single-cell level. The accuracy of the generated structures has been shown in our previous work (Shi and ThirumalaiPRX 2021). In ensemble-averaged Hi-C experiments, TADs appear to be relatively stable. However, imaging experiments (Bintu et. al, 2018) have revealed that TADs are not fixed structures present in every single cell, but instead exhibit variability at the single-cell level. TADlike structures with distinct boundaries are observed in individual cells, and the location of these boundaries varies from cell to cell. However, these TAD-like structures still show a preferential positioning in 3D structures. Interestingly, the preferential positioning often corresponds to TAD boundaries observed in population-averaged Hi-C data. This suggests that while cohesin is involved in establishing the overall organization of TADs, other factors and mechanisms could also contribute to TAD formation at the individual cell level. In this study, we showed some boundaries of P-TADs upon cohesin loss in the Hi-C maps, align with preferential boundaries in individual 3D structures of chromosomes. The makes the finding that a subset of TADs is preserved upon cohesin is robust.

From a technical perspective, the use of HIPPS avoids time-consuming polymer simulations. The HIPPS is rapid and can be used to generate arbitrarily large ensemble of structures, allowing us calculate properties both at the single cell and ensemble level.

In addition, we'd like to offer the following feedback to the authors.

1. The discussion of enhancer-promoter loops as a cause of TAD preservation is interesting, but it would be interesting to know fraction of preserved TADs enhancer-promoter loops might explain.

We thank the reviewer for the excellent suggestion. We have done the suggested calculation. The results are shown in a new Figure.8 in the main text. We also moved the results on enhancer-promoter to the main results section from the Discussion section.

1. The last paragraph of the introduction seems to state that only the HIPPS model was used to find single-cell 3D structures and boundary probabilities. However, the main text suggests that the CCM model was also used for these purposes.

We have revised the text to clarify this point on pages 3-4. Also please see the discussion on the utility of HIPPS above.

1. When referring to the boundary probability, it would be useful if the authors always specified whether they refer to the boundary probability before or after cohesin depletion (or loop depletion in the CCM model). Statements such as "This implies that peaks in the boundary probabilities should correspond to P-TADs" are ambiguous; it is unclear if the authors mean that boundary probabilities before cohesin depletion predict that the boundary will be preserved, rather than that preserved TAD boundaries correlate with peaks in the boundary probability after cohesin depletion.

We thank the reviewer for the suggestion. Indeed, it may be confusing. Hence, we have revised the text in numerous places to clarify this point.

1. It would be interesting to analyze all TAD boundaries that are present after cohesin depletion, rather than just those that overlap with TAD boundaries in WT cells. This would give better statistics for answering the question what causes TAD-like structures in cells without cohesin.

We thank the reviewer for this excellent suggestion. First, this would we believe this deviate from the primary goal of this study: what leads to TAD preservation after cohesin deletion? Second, this has to be done very systematically, as we did here for P-TADs, in order draw meaningful conclusions. This is a very useful study for another occasion.

1. The use of a plethora of acronyms (P-TAD, CM, DM, CCM, HLM...) makes the paper difficult to read.

We have revised the text to change CM to “contact map” and “DM” to “distance map”. For PTADs, CCM, and WLM, we would argue that P-TAD is rather a clear and intuitive abbreviation and CCM/WLM refers to specific methods/models and replacing them with full names would make text more difficult to read. We hope the reviewer is okay with us keeping these acronyms.

Reviewer #2 (Public Review):

Summary:

Here Jeong et al., use a combination of theoretical and experimental approaches to define molecular contexts that support specific chromatin conformations. They seek to define features that are associated with TADs that are retained after cohesin depletion (the authors refer to these TADs as P-TADs). They were motivated by differences between single cell data, which suggest that some TADs can be maintained in the absence of cohesin, whereas ensemble HiC data suggest complete loss of TADs. By reananalyzing a number of HiC datasets from different cell types, the authors observe that in ensemble methods, a significant subset of TADs are retained. They observe that P-TADs are associated with mismatches in epigenetic state across TAD boundaries. They further observe that "physical boundaries" are associated with P-TAD maintenance. Their structure/simulation based approach appears to be a powerful means to generate 3D structures from ensemble HiC data, and provide chromosome conformations that mimic the data from single-cell based experiments. Their results also challenge current dogma in the field about epigenetic state being more related to compartment formation rather than TAD boundaries. Their analysis is particularly important because limited amounts of imaging data are presently available for defining chromosome structure at the single-molecule level, however, vast amounts of HiC and ChIP-seq data are available. By using HiC data to generate high quality simulated structural data, they overcome this limitation. Overall, this manuscript is important for understanding chromosome organization, particularly for contacts that do not require cohesin for their maintenance, and for understanding how different levels of chromosome organization may be interconnected. I cannot comment on the validity of the provided simulation methods and hope that another reviewer is qualified to do this.

We appreciate the reviewer for a comprehensive summary of our work, and we are happy that the reviewer finds our work important, which provides valuable insights to the field.

Specific comments

It is unclear what defines a physical barrier. From reading the text and the methods, it is not entirely clear to me how the authors have designated sites of physical barriers. It may help to define this on pg 7, second to last paragraph, when the authors first describe instances of PTAD maintenance in the absence of epigenetic mismatch.

We thank the reviewer for the suggestions. The details of physical boundary designation are provided in the appendix data analysis. To make the concept and idea of physical boundary easy to understand, we have revised the text on page 7 in the revised main text.

Figure 7 adds an interesting take to their approach. Here the authors use microC data to analyze promoter-enhancer/promoter-promoter contacts. These data are included as part of the discussion. I think this data could be incorporated into the main text, particularly because it provides a biological context where P-TADs would have a rather critical role.

We thank the reviewers for the suggestion. We also agree that results in Figure 7 provide novel insights on TAD formation and its possible preservation upon perturbation. We have followed the reviewer’s suggestion to move it to an independent section in the main results section as the last subsection.

Figure 3a- the numbers here do not match the text (page 6, second to last paragraph). The numbers have been flipped for either chromosome 10 or chromosome 13 in the text or the figures.

We thank the reviewer for pointing out this error. In the revised main text, it has been corrected.

Reviewer #3 (Public Review):

This manuscript presents a comprehensive investigation into the mechanisms that explain the presence of TADs (P-TADs) in cells where cohesin has been removed. In particular, to study TADs in wildtype and cohesin depleted cells, the authors use a combination of polymer simulations to predict whole chromosome structures de novo and from Hi-C data. Interestingly, they find that those TADs that survive cohesin removal contain a switch in epigenetic marks (from compartment A to B or B to A) at the boundary. Additionally, they find that the P-TADs are needed to retain enhancer-promoter and promoter-promoter interactions.

Overall, the study is well-executed, and the evidence found provides interesting insights into genome folding and interpretations of conflicting results on the role of cohesin on TAD formation.

We are pleased with the reviewer’s positive assessment of our work.

To strengthen their claims, the authors should compare their de-novo prediction approach to their data-driven predictions at the single cell level.

We thank the reviewer for the very good suggestion. We are assuming that the Reviewer is asking us to compare the CCM simulations with HIPPS generated structures at the single cell level. We have shown, using the GM12878 cell data, that the polymer simulations reproduce the Hi-C contact maps (an average quantity) well (see Appendix Fig. 2 and Fig. 3). In addition, we show in Appendix Fig. 8 the comparison with ensemble averaged distance maps as well as at the single cell level for Chr 13 from the GM12878 cell. There are TAD-like structures at the single cell level just as we find for HCT-116 cell (Fig. 5 in the main text). Thus, the conclusions from de-novo prediction and data-driven predictions are consistent. In addition, in our previous publication introducing HIPPS in Phys Rev X 11: 011051 (2021), we showed that the method is quantitatively accurate in reproducing experimental data for all the interphase chromosomes.

Having demonstrated this consistency, we used computationally simple data-driven predictions to analyze HCT-116 and mouse liver cell lines for which Hi-C data with and without cohesin rather than perform multiple laborious polymer simulations.

Please see below for our response to specific comments.

1. It is confusing that the authors change continuously their label for describing B-A and A-B switches. They should choose one expression. I think that the label "switch" between A and B is more precise than "mismatch".

We have revised the text to make it consistent. Now it all reads “A-B”. Yes, the suggestion that we use switch is good but we think that mismatch is more concise. We trust that this Reviewer will indulge us on this point.

1. In the Abstract, the authors mention HCT-116 cells but do not specify which cells are these.

We have changed “HCT-116” in the abstract to “human colorectal carcinoma cell line”.

1. In the Abstract, it is unclear what the authors mean by "without any parameters"

In the theoretically based HIPPS method, there is no “free” parameter. In other words, the only parameter is uniquely determined. To avoid confusion, we have removed “without any parameters” from abstract.

1. In Results, what do the authors mean by 16% (26%)?

This refers the percentage of how many TADs are preserved after Nipbl and RAD21 removal in mouse and HCT-116 cells, respectively. Using TopDom method, we identified TAD boundaries in Wild and cohesin-depleted cells. There are 16% (959 out of 4176 – Fig. 1a) and 26% (1266 out of 4733 – Fig. 1b) of TADs are preserved after Nipbl and RAD21 removal in mouse and HCT-116 cells, respectively. We removed the percentages in the revised version.

1. In Results, the authors mention "more importantly, we did tune the value of any parameter to fit the experimental CMs". Did they mean that instead they didn't tune any parameter?

We apologize for the confusion. In the CCM, there is a single controlled parameter. We have changed the sentence to reflect this correctly.

1. In Results, section "CCM simulations reproduce wild-type Hi-C maps", Kullback-Leibler (KL) divergence is used to assess the correlation between two loci, but it is unclear what the value 0.04 stands for; is it a good or a bad correlation?

The value for Kullback-Leibler divergence can vary from 0 to infinity with 0 give the perfect correlation. Thus, 0.04 means that the correlation is excellent.

1. The authors use two techniques to obtain 3D structures, one is CCM, which takes the cohesin as constraints, and another is HIPPS, which reconstructs from Hi-C maps. Both seem to have good agreement with the Hi-C contact maps. However, did the authors compare the CCM with the HIPPS 3D structures?

This is detailed in response at the start of the reply to this Reviewer. As detailed in this response as well in the main text we used the CCM to generate hypotheses for the origin of P-TADs. In the process, we established the accuracy of CCM, which gives us confidence about the hypotheses. As explained above and emphasized in the revised version, CCM simulations are time consuming whereas generating 3D structures using HIPPS is computationally simple. Because HIPPS is also accurate, we used it to analyze the Hi-C data on mouse liver, HCT-116 as well as Micro-Data on mESC.

In our paper in Phys Rev X 11: 011051 (2021) we showed that HIPPS reproduces Hi-C data. In the current manuscript, we showed in Appendix Fig. 2 and Fig. 3 as well as in a study in 2018 (Shi and Thirumalai, Nat Comm.) that CCM is accurate as well. Thus, there is little doubt about the accuracies of the methods that we have developed.

1. In Results, section "P-TADs have prominent spatial domain boundaries", the authors constructed individual spatial distance matrices (DMs) using 10,000 simulated 3D structures. What are the differences among these 10,000 simulations? Do they start them with different initial structures?

The structures are generated using HIPPS which is data-driven method that uses Hi-C contact map as constraints. The method, which uses the maximum entropy theory, samples from a distribution that describe the structural ensemble of chromosome. The 10,000 structures are randomly sampled and are independent from each other. The HIPPS method is not a simulation, and hence the issue of initial structures does not arise.

1. In Methods, when the authors mention the "unknown parameter", do they use one parameter for all simulations (+/- cohesin) or is this parameter different for each system? Would this change the results?

We apologize for the confusion. The “unknown parameter” is the energy scale 𝜖 that describes the interaction strength between chromosome loci. We have revised the text in the method (page 27) to clarify it. The same value of 𝜖 is used for all CCM simulation with or without cohesin.

1. In Methods, when the authors perform DBSCAN clustering, they mention that they optimize the clustering parameters for each system. However, if they want to compare between different systems, the clustering parameters should be the same.

The purpose of DBSCAN is to capture the spatial clustering topology of chromosome loci. However, different cell types and chromosomes may have different overall density, which will impact the average distance between loci. If using the same parameters, such global changes will impact the result of clustering most and the intended spatial clustering topology can be distorted. Hence, we tune the clustering parameter for each system in order to ignore the global effect but only capture the local and topology of clustering of chromosome loci.

Grammar comments:

1. "structures, with sharp boundaries are present, at.."

We thank the reviewer for pointing out the error. We have fixed it.

1. "Three headlines emerge from these studies are:"

We have fixed it.

1. "both the cell lines"

We have fixed it.

Version 2

Abstract

Compartment formation in interphase chromosomes is a result of spatial segregation between eu- and heterochromatin on a few mega base pairs (Mbp) scale. On the sub-Mbp scales, Topologically Associating Domains (TADs) appear as interacting domains along the diagonal in the ensemble averaged Hi-C contact map. Hi-C experiments showed that most of the TADs vanish upon deleting cohesin, while the compartment structure is maintained, and perhaps even enhanced. However, closer inspection of the data reveals that a non-negligible fraction of TADs is preserved (P-TADs) after cohesin loss. Imaging experiments show that, at the single-cell level, TAD-like structures are present even without cohesin. To provide a structural basis for these findings, we first used polymer simulations to show that certain TADs with epigenetic switches across their boundaries survive after depletion of loops. More importantly, the three-dimensional structures show that many of the P-TADs have sharp physical boundaries. Informed by the simulations, we analyzed the Hi-C maps (with and without cohesin) in mouse liver and human colorectal carcinoma cell lines, which affirmed that epigenetic switches and physical boundaries (calculated using the predicted 3D structures using the data-driven HIPPS method that uses Hi-C as the input) explain the origin of the P-TADs. Single-cell structures display TAD-like features in the absence of cohesin that are remarkably similar to the findings in imaging experiments. Some P-TADs, with physical boundaries, are relevant to the retention of enhancer-promoter/promoter-promoter interactions. Overall, our study shows that preservation of a subset of TADs upon removing cohesin is a robust phenomenon that is valid across multiple cell lines.

Introduction

Advances in experimental techniques have provided glimpses of the three-dimensional (3D) organization of chromosomes in diverse species[1–6]. The average (performed over a large number of cells) contact map [4, 6], inferred using chromosome conformation capture technique and related variants (referred to as Hi-C from now on), is a two-dimensional (2D) matrix, whose elements are a measure of the probability that two loci separated by a certain genomic distance are spatially adjacent. The Hi-C experiments on different mammalian cells suggest that there are two major length scales in the organization of interphase chromosomes. On the scale, LC ∼ (2-5) Mbp (mega base pairs or Mbp), one observes checkerboard patterns in the contact maps [4], which are thought to be associated with micro-phase separation between the two major epigenetic states, active (A) or euchromatin and inactive (B) or heterochromatin. On the length scale, LTAD, from tens of kb up to a few Mb, domains, referred to as Topologically Associating Domains (TADs), appear as squares along the diagonal of the contact maps [7, 8]. Contacts are enriched within the TADs, and are suppressed across the boundaries between the TADs. A number of polymer models [9–14] have shown that compartment formations and TADs may be explained using micro-phase separation between A and B type loci. The use of two length scales, LC and LTAD, in characterizing the organization of interphase chromosomes is now entrenched in the field, although there are suggestions that finer sub-TAD structures emerge at kilo base scales [15, 16]. In particular, recent Micro-C experiments have shown that there are fine structures starting from the nucleosome level [17–19], thus establishing the hierarchical organization of interphase chromosomes over a broad range of length scales.

TADs are thought to regulate gene expression by constraining the contacts between target gene and regulatory regions [20, 21]. As a consequence, perturbation or disruption of their integrity such as deletions, duplications or inversions of DNA segments within the TADs, could lead to aberrant gene expression [8, 22–28]. A class of chromatin loops, mediated by the ATP-dependent motor cohesin [29] and the DNA-binding protein CTCF protein (“cohesin-associated CTCF loop”), organizes a subset of the TADs [30]. It is thought that cohesin [29] extrudes DNA loops of varying lengths, which are terminated when the motor encounters the transcriptional insulator CCCTC-binding factor (CTCF) [31]. This implies that cohesin and CTCF are often colocalized at the TAD boundary [4–6, 30, 32, 33].

Several experiments have shown that depletion of the architectural proteins (Nipbl, RAD21, and CTCF) disrupts the organization of interphase chromosomes [26, 34–41].

Schwarzer et al.[34] showed that the removal of the cohesin loading factor, Nipbl in the mouse liver cell results in loss of TADs. They concluded that compartment formation, which is independent of cohesin, is a consequence of the underlying epigenetic landscape, while TAD formation requires cohesin. Similarly, it was found that upon removal of cohesin subunit, RAD21 cohesin-associated CTCF loops and TADs are abolished [26, 39, 40]. Deletion of RAD21 results in the complete loss of the so-called loop domains [26], which are formed when CTCF colocalizes with cohesin. In contrast, imaging experiments[40] showed that TAD-like structures, with sharp boundaries, at the single-cell level survive even after deleting cohesin. Three headlines emerged from these studies: (i) They reinforce the two-length scale description of genome organization at the ensemble level. (ii) Factors that prevent the association of cohesin with chromosomes globally abolish the TADs and the Hi-C peaks, but preserve (or even enhance) compartmentalization. Experimental studies [4, 26, 34, 39, 40] and polymer simulations [12, 35, 42] have shown that the global epigenetic state determines compartment formation, while the more dynamic TADs, with finite lifetimes[43], require ATP-dependent cohesin. (iii) TAD-like features persist in single cells before and after auxin treatment, albeit with changes in the locations of the sharp domain boundaries.

The results of super resolution experiments [40], at the single-cell level (described above), made us wonder if there is evidence for preservation of TADs at the ensemble level upon cohesin depletion. To this end, we first analyzed the experimental contact maps from mouse liver and HCT-116 cells (human colorectal carcinoma cell line), in the presence and absence of cohesin to assess if TADs are preserved. We discovered that, on an average a fraction of TADs, identified using the TopDom method [44], are retained in chromosomes from both the cell lines (Fig. 1) after removing cohesin. These findings raise the following questions. What is the mechanistic basis for the retention of a small but significant fraction of TADs that are preserved after cohesin loss? Is there a structural explanation for TAD retention at the ensemble level (Hi-C), which would reconcile with the results in super resolution imaging experiments showing TAD-like structures at the single level, even without cohesin?

Fate of the TADs in chromosomes upon cohesin deletion.

(a) The number of TADs in all the chromosomes, identified by the TopDom method [44], in the wild type (WT) cells and the number of preserved TADs (P-TADs) after deleting cohesin loading factor (Nipbl) in mouse liver. (b) Same as (a) except the experimental data are analyzed for HCT-116 cell before (WT) and after RAD21 deletion. (c) The total number of TADs and the number of P-TADs for each chromosome calculated using the mouse liver Hi-C data. The number above each bar is the percentage of P-TADs in each chromosome. (d) Same as (c) except the results are for chromosomes from the HCT-116 cell line. The percentage of P-TADs is greater in the HCT-116 cell line than in mouse liver for almost all the chromosomes, a feature that is more prominent in the distribution of P-TADs proportions (Right).

We answered the questions posed above by using the following strategy. We first performed polymer simulations for two chromosomes from the GM12878 cell line using the Chromosome Copolymer Model (CCM) [11] with and without loop anchors, which mimics the wild type (WT) and the absence of cohesin-associated CTCF loops, respectively. The major purpose of the CCM polymer simulations is to determine the mechanisms for the emergence of P-TADs. Because the simulations directly generate 3D structures, they can be used to compute average contact maps that can be compared to experiments to determine the accuracy of the CCM. In addition, comparisons of the contact maps with and without cohesin, allowed us to generate the mechanisms for the emergence of P-TADs. Using the polymer simulations of chromosomes from the GM12878 cell line [4], whose organization without cohesin is unknown, we determined that P-TADs arise due to epigenetic switches across TAD boundaries and/or associated with peaks in boundary probabilities, which require knowledge of ensemble of 3D structures.

Informed by the results from the polymer simulations, we analyzed the experimental data from two cell lines [26, 34]. We discovered that epigenetic switch does account for a reasonable fraction (≈ 0.4 in mouse liver and ≈ 0.3 in HCT-116 cell lines) of P-TADs. Rather than perform multiple time-consuming polymer simulations, we generated the 3D structural ensemble using the accurate and data-driven HIPPS method [45] of the chromosomes using the experimental Hi-C data. The analyses using the 3D structures accounted for about 53% of the P-TADs, predicted by theTopDom method [44]. Strikingly, the 3D structures revealed TAD-like structures, at the single-cell level, both in the presence and absence of cohesin, which is in accord with the super resolution imaging data [40]. Our work shows that the effects of cohesin removal on chromatin structures are nuanced, requiring analyses of both the epigenetic landscape as well as 3D structures in order to obtain a comprehensive picture of how distinct factors determine interphase chromosome organization in the nucleus. Our calculations for chromosomes from three cell lines lead to the robust conclusion that a subset of P-TADs is intact after deleting cohesin.

Results

A non-negligible fraction of TADs are preserved upon removal of cohesin

Experiments [26, 34–37, 39, 40] have shown that deletion of cohesin loaders (Nipbl in mouse liver, SCC2 in yeast), cohesin subunit (RAD21) abolishes a substantial fraction of both cohesin-associated CTCF loops and the TADs. These observations across different cell lines raise an important question: Do all the TADs completely lose their contact patterns after removal of cohesin? To answer this question, we first analyzed 50kb-resolution contact maps from the two cell lines (mouse liver [34] and HCT-116 [26]) before and after degradation of Nipbl and RAD21, respectively (see Appendix Analyses of the experimental data for details). Using TopDom [44], we discovered that roughly 659 TADs out of 4176 (16%) are preserved (Fig. 1a) after removing Nipbl in the mouse liver cells. In the HCT-116 cells, 1226 TADs out of 4733 (26%)) are preserved (Fig. 1b) upon RAD21 loss. Fig. 1c and Fig. 1d show that the number of P-TADs depends on the chromosome number. Although the actual number of P-TADs would depend on the TAD-calling protocol, the the finding that a non-negligible fraction is preserved after cohesin depletion is highly significant.

CCM simulations reproduce wild-type Hi-C maps

To explore the mechanism resulting in P-TADs, we first simulated the chromosome copolymer model [11], CCM (Appendix Fig. 2a). To independently decipher the origins of P-TADs (Fig. 1) in experiments, we calculated the contact maps for Chr13, shown in Appendix Fig. 2 (Chr10 in Appendix Fig. 3,) from the GM12878 cell line. The CCM simulations (Appendix Fig. 2b) reproduce the ubiquitous checker board patterns well. The rectangle in Appendix Fig. 2b represents the border of one such compartment formed primarily by interactions between the B-type loci.

Identification of P-TADs from the contact map using the TopDom method.

(a) Schematic representation used to determine the P-TADs. Yellow (Blue) triangles represent the TADs identified using TopDom method in WT (cohesin depleted) contact maps at 50kb resolution.Small square within each triangle represents a single locus (50kb size). The boundaries of a TAD detected in the WT contact map within ± one bin (50kb) from a position of boundaries in cohesin depleted cells is deemed to be a P-TAD. (b) P-TAD upon cohesin loss in HCT116 cell. The bar plots above the contact maps show the epigenetic states. Red (Blue) color represents the active (inactive) state, respectively. The TAD between grey dashed lines is preserved upon cohesin loss. The parameter (with red square) displayed at each left bottom indicates the color scale when plotting contact maps used in Juicebox[54].

CCM simulations reveal characteristics of P-TADs.

(a) The number of TADs in the simulated Chr10 and Chr13 chromosomes for PL=1. The number of P-TADs after CTCF loop depletion (PL=0) is also shown. (b) The number of P-TAD with epigenetic switches (blue) and those identified by the peaks in the boundary probability (green). (c)-(e) Comparison between contact maps for the region of Chr13 with upper (lower) triangle with PL = 1 (PL = 0). The black circles at the corner of the TADs are the CTCF loop anchors. The bar above the contact map are the epigenetic states with red (blue) representing A (B) loci. Arrows above the bar show the epigenetic switch. (c) After loop deletion, TAD structures disappear. (d) TAD whose boundaries are marked by epigenetic switches are preserved. (e) TAD lacking at least one epigenetic switch is disrupted after loop loss. (f)-(h) Comparison of the contact map and the mean spatial-distance matrices for the 2.5Mb genomic regions (25.7-28.2Mbp,73.3-75.8Mbp and 102-104.5Mbp, respectively) with (upper) and without (lower) loop anchors. Bottom graph shows the boundary probability, with the high values indicating population averaged TAD boundary. Purple circles in the boundary probability graph represent the preferred boundaries. A subset of P-TADs boundaries match epigenetic switches (blue lines). P-TADs with high boundary probability is in green line. The magenta line describes P-TADs which is not accounted by epigenetic switch or physical boundary in 3D space but are found using the TopDom method.

In order to quantitatively compare the Hi-C data and the simulated contact maps, we transformed the contact maps to Pearson correlation maps, which is used to assess if two loci have correlated interaction profiles (Appendix Fig. 2c). The Kullback-Leibler (KL) divergence between the two probability distributions for the Pearson correlation coefficients, ρijs, from simulations and experiments is 0.04 (see Appendix Fig. 2d). We also performed Principal Component Analysis (PCA) on the Pearson correlation matrix to identify the compartment structure. Comparison of the PCA-derived first principal components (PC1) across the Chr13 reveals that A/B compartments observed in the CCM correspond well to those found in the experiments (Appendix Fig. 2e).

We then compared the three-dimensional spatial organization between the simulations and experiments using the Ward Linkage Matrix (WLM), which is based on an agglomerative clustering algorithm. The simulated WLM is calculated from a spatial distance map of the organized chromosome (described in Appendix Fig. 2f-h). We constructed the experimental WLM by converting the Hi-C contact map to a distance map using the approximate relationship [2, 11, 45],. Here, Pij is the contact probability, and Rij is the 3D spatial distance between loci i and j (see Appendix Data Analyses). The Pearson correlation coefficient between experimental and simulated WLMs is 0.83 (Appendix Fig. 2h), which establishes the accuracy of the CCM.

Snapshots of TAD structures in Appendix Fig. 2k and n show that they are compact but structurally diverse. The average sizes of the TADs detected using TopDom [44] from Hi-C and simulated contact maps are ∼ 615kbs and ∼ 565kbs, respectively. Overall the emerging picture of the compartment and TAD structures using different methods are consistent with each other. The results in Appendix Fig. 2 show that the agreement between the CCM simulations and Hi-C data is excellent, especially considering that (i) error estimates in the Hi-C experiments are essentially unknown, and (ii) more importantly, only a single parameter, the inter-loci interaction strength, ϵ, is tuned to fit the experimental contact maps (see Appendix methods sections). Taken together, the results show that the key features of the Hi-C maps for Chr13 (see Appendix Fig. 3 for Chr10 results) are accurately reproduced by the CCM simulations.

Epigenetic switch accounts for a large fraction of P-TADs

Most of the TADs are not discernible after loop loss, as evidenced by the blurred edges in the contact maps (Fig. 3c). In the CCM simulations of chromosomes from the human GM12878 cell line, a subset of TADs remains even after deleting cohesin-associated CTCF loops (Fig. 3a). The percentages of P-TADs depend on both the resolution of the Hi-C experiments as well as the algorithm used to identify the TADs. By using the same method to analyze both the simulation results and experimental data, it is hoped that the conclusions would be robust.

We used the simulation results to determine the mechanism for the emergence of P-TADs by comparing the results for PL=1 and PL=0. The first observation is that some TADs, even with cohesin-associated CTCF loops, consist mostly of sequences in the same epigenetic state (A or B). Fig. 3d compares the fate of one such TAD in the region (19.3 -21.8 Mb) in Chr13 between PL=1 and PL=0. The highlighted TAD is preserved upon loop loss although the probabilities of contact within this TAD is reduced when PL = 0 (bottom) compared to PL = 1 (top). Their boundaries correspond to a switch in the epigenetic state (the sequence location where the change in the epigenetic states occurs, as shown by the two black arrows). In contrast, a TAD in Fig. 3e, that is present in the WT, is abolished when PL=0. The disruption of this particular TAD, lacking in at least one epigenetic switch, occurs because it can interact more frequently with neighboring TADs composed of similar epigenetic states, which in this case is B-type loci. The importance of epigenetic switch in the P-TADs has been noted before [26].

The results in Figs. 3d and e show that a switch in the epigenetic state across a TAD boundary in the WT is likely to result in its preservation after cohesin-associated CTCF loop loss. To test this assertion, we calculated the number of P-TADs that are associated with switches in the epigenetic states (see Appendix Data Analyses and Appendix Fig. 1 for details). We considered the TAD boundary and epigenetic switch as overlapping if they are less than 100kb apart, which is reasonable given that Hi-C resolution adopted here is 50kb. By using 100kb as the cut off, we only consider switch occurrences that exceed two loci. With this criterion, out of 216 (169) TADs calculated using TopDom 50 (23) are P-TADs for Chr10 (Chr13) (vertical blue bars in Fig. 3b in which there are epigenetic switches in the WT).

The P-TADs with epigenetic switches, illustrated in Fig. 3f, shows TADs in the 2.5Mbs region in Chr13. Among the three P-TADs, two of them, whose boundaries are marked by dashed blue lines, have epigenetic switch across the TAD boundary. These two TADs survive after removal of cohesin-associated CTCF loop.

P-TADs have prominent spatial domain boundaries

Because there are a number of P-TADs that are preserved even without epigenetic switches across their boundaries, we wondered if the distance matrix, which requires 3D structures, would offer additional insights about P-TADs upon cohesin-associated CTCF loop loss. Recent imaging experiments[40, 42, 46] revealed that TAD-like domain structures with spatially segregated heterogeneous conformations are present in single cells even without cohesin. The physical boundaries of TAD-like domains, identified from individual 3D structures, vary from cell to cell. They exhibit a preference to reside at specific genomic positions only upon averaging, as found in the Hi-C experiments. The boundary probability at each locus is the proportion of chromosome structures in which the locus is identified as a domain boundary in the 3D space. The locations of prominent peaks in the boundary probability frequently overlap with TADs detected by the population level Hi-C maps.

To explore the relation between P-TADs in ensemble averaged contact maps and preferential boundaries in individual 3D structures of chromosomes, we first calculated individual spatial distance matrices using 10,000 simulated 3D structures which were used to identify the single-cell domain physical boundaries [40]. The physical domain boundaries identified from the 3D structures are the chromosome loci that spatially separate two physical clusters. It is constructed by comparing the spatial distances between a reference locus with the up and down-stream chromosome segments [40]. Specifically, we calculated the median values of pairwise distances between the reference loci and the up-stream loci, and also the median values of pairwise distances between the reference loci and the down-stream. The ratio of these two quantities is defined as boundary strength. If a locus’s boundary strength is above a predefined threshold, this locus is defined as a physical boundary locus. The idea is that a physical boundary has large ratio, as it spatially separates up-stream and down-stream chromatin segments. Based on these boundary positions in individual cells, we define the boundary probability of a locus as the probability (fraction of all individual structures) of this locus being the physical boundary in an ensemble of individual structures. The detailed mathematical definition is provided in Appendix Data Analyses and illustrated in Appendix Fig. 6. We find preferential domains, with high peaks, in the boundary probability along the genomic region as well as variations in single-cell domains both in PL = 1 and PL = 0 (see Appendix Fig. 8).

The CCM simulations show that TADs with epigenetic switches across the boundary are likely to be preserved after cohesin-associated CTCF loop loss. Furthermore, Fig. 3f (blue dashed lines) shows that single-cell domain boundaries preferentially reside at the TAD boundaries with epigenetic switches, leading to a prominent signature for the structural ensemble after averaging over a population of cells. Interestingly, the P-TAD has prominent peaks in the boundary probabilities (in both the WT and cohesin-depleted cells), sometimes even without epigenetic switch, at the same genomic position as in the contact map (green lines in Fig. 3f and g). These observations imply that the presence of physical boundaries in the 3D structures may be used to identify P-TADs, especially in instances when there are no epigenetic switches. We should note that the simultaneous presence of peaks in the boundary probabilities in the both WT and cohesin-depleted cells is a signature of P-TADs.

Not all preferential boundaries identified in the distance matrices of the WT cells coincide with the TADs detected using the contact map [40]. There is discordance in the TAD boundaries and high peaks in the boundary probability. The top panel in Fig. 3h (magenta lines) shows that in the (102 - 104.5) Mb range, TopDom predicts that there are three P-TADs after loop loss (see the top panel with PL = 0). There are two prominent peaks in the WT boundary probability whose boundaries coincide with the TADs predicted by TopDom (see bottom panel with PL=1 in Fig. 3h). But the peak height for the third TAD is very small. At best, one can deduce from the boundary probabilities (compare the results in Fig. 3h for PL=1 and PL=0) that the middle TAD is preserved, which would be consistent with the TopDom prediction.

We calculated a standardized Z-score for the boundary probability in the genomic region in order to determine the preferred boundaries in single-cell domains. The number of P-TADs that are accounted for by prominent boundary peaks increases if Z-score is reduced. This implies that some P-TADs detected in the contact maps using TopDom have weak physical boundaries in the 3D structures. We considered the maxima, with Z-score values larger than 0.7, as preferred boundaries in order to determine if P-TADs arise due to the presence of strong physical boundary. With this criterion, we obtained good agreement for the mean length of the TADs detected in the contact map using the TopDom method. The averaged sizes of the TADs in Chr13 using TopDom and boundary probability are ∼565kbs and ∼535kbs, respectively. Quantitative analysis of the boundary probabilities along the genomic region revealed ≈ 66% of the P-TADs in Chr10 and Chr13 have preferential positioning in single-cell domains (green bars in Fig. 3b). Most P-TADs with epigenetic switches display prominent peaks in the boundary probabilities (≈ 85%).

The primary lessons from the simulations, which form the basis for analyzing the experiments on chromosomes from mouse liver and HCT-116 cell lines, are: (i) switch in the epigenetic state across the TAD boundary is a predominant factor in determining the P-TADs after CTCF loop deletion. (ii) The presence of peaks in the boundary probabilities in both the WT and cohesin-depleted cells, calculated from the the 3D structures, accounts for certain fraction of P-TADs. However, in some instances TopDom predictions (used in (i)) are not compatible with boundaries deduced from 3D structures. (iii) The polymer simulations show that the ensemble of 3D structures provides insights into the consequences, both at the single cell and ensemble averaged level, of depleting cohesin.

Structural explanation of P-TADs upon cohesin removal from analysis of Hi-C data

In order to assess if the conclusions from simulations, summarized above, explain the experimental data, we first calculated the number of P-TADs whose boundaries have switches in the A/B epigenetic states in mouse liver and HCT-116 cell lines. We assigned chromatin state A (active, red) or B (repressive, blue) by analyzing the combinatorial patterns of histone marks using ChromHMM [47] (see Appendix analyses of the experimental data and Appendix Fig. 7). An average over the 20 chromosomes shows that 280 P-TADs were associated with a switch between A and B epigenetic states upon Δ Nipbl in mouse liver (blue bar in Fig. 4a). The corresponding number of P-TADs, averaged over 23 chromosomes, with epigenetic switches is 396 after deleting RAD21 in HCT-116 (blue bar in Fig. 4b). Not unexpectedly, TADs with epigenetic switches across their boundaries are preserved with high probability after cohesin deletion. We also find that a large number of P-TADs are accounted (green bars in Fig. 4a and Fig. 4b) for by the presence of peaks in the boundary probabilities in the cohesin-depleted cells, which we discuss in detail below.

Classification of P-TADs from Hi-C maps from two cell lines and link between boundary probability peak and epigenetic switch.

(a) The number of P-TADs in all the chromosomes (orange bar taken from Fig. 1a) that are accounted for by epigenetic switches (blue bar) as well as peaks in the boundary probability (green bar) after Nipbl loss in mouse liver. (b) Same as (a) except the analyses is done using experimental data are for HCT-116 cell after RAD21. (c) Example of P-TAD in the WT 97.7Mb-100.2Mb region of Chr3 from HCT-116 cell line. The mean distance matrices calculated using the 3D structures is in the middle panel.The dark-red circles at the boundaries of the TADs in the contact maps are loop anchors detected using HiCCUPS [55]. The peaks in the boundary probability (bottom panel) are shown by purple circles. Epigenetic switch coincides with peak in the boundary probability (compare top and bottom panels). Bottom plot shows the probability for each genomic position to be a single-cell domain boundary. (d) Same as (c) except the results correspond to the absence of RAD21. Although not as sharp, there is discernible peak in the boundary probability when there is an epigenetic switch after removal of RAD21.

We then searched for a structural explanation for P-TADs in the two cell lines (Fig. 4a and Fig. 4b). A plausible hint comes from the the CCM simulations (Fig. 3), which show that boundary probabilities, whose calculations require 3D structures, are good predictors of P-TADs. This implies that peaks in the boundary probabilities should correspond to P-TADs. Similar findings are obtained by analyzing the experimental data (Fig. 4c and d). Additional examples are discussed in Fig. 5. In light of these findings, we wondered if, in general, physical boundaries can be inferred directly using Hi-C data from ensemble experiments, instead of performing tedious polymer simulations. To this end, we used the HIPPS method (see Methods), to calculate an ensemble of 3D structures with the Hi-C contact map as the only input. Several conclusions follow from the results in Fig. 5. (i) Figs. 5a and b show that HIPPS faithfully reproduces the Hi-C contact maps. (ii), Using the ensemble of 3D structures, we calculated the locus-dependent boundary probabilities for both the WT and cohesin-depleted cells. Comparison of the peak positions in the averaged boundary probabilities and the TAD boundaries shows that they often coincide, although there are discordances as well (Fig. 5 and Appendix Figs.12 and 13). (iii) When there is a switch in the epigenetic states, a substantial fraction of P-TADs have high peaks in the boundary probabilities (see Fig. 5 and Appendix Fig. 12). As in the simulations, a large fraction of P-TADs (≈ 67%) have high peaks in the boundary probabilities. Taken together, the results show that the predictions using boundary probabilities and the TopDom method are consistent. (iv) Analyses of the experiments suggest that the epigenetic state as well as the presence of physical boundary in the 3D structures have to be combined in order to determine the origin of P-TADs in cohesin depleted cells.

Fate of TADs after ΔNipbl in mouse liver cells.

(a)-(b) Comparison between Hi-C (lower) and calculated contact maps (upper) using the 3D structures obtained from HIPPS method for the 3Mb genomic regions (Chr6:22.6Mb-26.1Mb in WT cells and Chr7:139Mb-142.5Mb in Nipbl -depleted cells), respectively. The distance threshold for contact is adjusted to achieve the best agreement between HIPPS and experiments. Calculated contact maps are in very good agreement with Hi-C data for both WT and Nipbl -depleted cells. (c) Complete Loss (Chr6:23.55Mb-26.05Mb) of TADs in ΔNipbl. (d)-(e) P-TADs (Chr7:139.5Mb-142Mb and Chr15:89.5Mb-92Mb). The plots below the scale on top, identifying the epigenetic states[47], compare 50kb-resolution Hi-C contact maps for the genomic regions of interest with Nipbl (upper) and without Nipbl (lower). Mean spatial-distance matrices, obtained from the Hi-C contact matrices using the HIPPS method [45], are below the contact maps. The dark-red circles at the boundaries of the TADs in the contact maps are loop anchors detected using HiCCUPS [4]. ChIP-seq tracks for CTCF, RAD21 and SMC3 in the WT cells [34] illustrate the correspondence between the locations of the most detected loop anchors and the ChIP-seq signals. Bottom plots give the probabilities that each genomic position is at a single-cell domain boundary in the specified regions. Purple circles in the boundary probability graph represent the physical boundaries. A subset of physical boundaries in P-TADs coincide with epigenetic switches (blue lines), indicating that the probabilities of contact at these boundaries are small. P-TADs in (e), demarcated by green lines, have high peaks in the boundary probability in the absence of epigenetic switch.

Certain TADs enrichhed in E-P/P-P interactions at the boundary are preserved pon cohesin deletion.

(a) Comparison between 5kb Micro-C contact maps in the region (Chr8:72.24Mb-72.57Mb) for the WT (left panel) and cohesin-depleted (right panel) mESC cells[48]. Location of cohesin loops (green square) and EP/P-P (blue circles) plotted in the WT contact maps are from experiments[48]. Bars above the contact map show epigenetic states (Red: Active, Blue: Inactive) annotated based on ChromHMM results[56]. The cohesin-dependent (green dashed lines) and independent (blue dashed lines) TADs are detected in the WT cells using the TopDom method with default parameter (w=5). P-TADs (blue dashed lines) are also found in cohesin deleted cells. (b)-(c) Comparison between 20kb Micro-C contact maps and mean distance maps spanning the regions, Chr19:8.66-9.2Mb and Chr12:56.4-56.9Mb, respectively, in the presence (upper) and absence (lower) cohesin. Bottom graph, below the distance maps, shows the boundary probability calculated from 10,000 3D structures. P-TADs between grey dashed lines are detected using TopDom method (w=5). A P-TAD with high boundary peak, without epigenetic switches, are enriched due to E-P/P-P interactions at the boundaries.

Statistics of the TADs in chromosomes upon cohesin loss using Micro-C contact data. The number of TADs in all the chromosomes in the wild type (WT, dark blue bar), the total number of preserved TADs (P-TADs, light blue bar) after deleting RAD21, and the number of P-TADs whose boundaries coincide withe enhancer-promoter/promoter-promoter (E-P/P-P) interactions (magenta bar) in mESC. About a third of the P-TADs are associated with E-P/P-P interactions.

With the near quantitative agreement with experiments, we performed detailed analyses, based on the epigenetic switches and boundary probabilities for chromosomes from the mouse liver (Fig. 5). The Appendix contains analyses of the experimental data, and the results for HCT-116 cell are given in Appendix Fig. 12. To illustrate different scenarios, we consider the 2.5Mbs regions from Chr6 (Fig. 5c), Chr7 (Fig. 5d), and Chr15 (Fig. 5e). (i) For Chr 6, there are three TADs according to TopDom (Fig. 5c) in the WT. Upon ΔNipbl, these TADs are abolished (compare the top and bottom panels in Fig 5c). The epigenetic track indicates that the region is mostly in the repressive (B) state. Quantification of the boundary probabilities along the 2.5Mb region of Chr6 shows that the TADs also lack physical boundaries upon ΔNipbl. (ii) Examples of P-TADs that satisfy the epigenetic switch criterion are given in Fig. 5d. Using TopDom, we identified several TADs (top panel in Fig. 5d) in this region of Chr7. It is interesting that the boundary probabilities obtained from the HIPPS-generated distance matrices are also large when there is a switch in the epigenetic states. In these examples, both epigenetic switches and boundary probabilities give consistent results (see the dashed blue lines in Fig. 5d). Two TADs in the WT (the ones on the right in the upper panel in Fig. 5d) merge to form a single TAD in the Δ Nipbl. This observation is in accord with expectation based on epigenetic switch, whose corollary is that if there is a TAD within a region that contains predominantly A or B type loci they ought to merge upon ΔNipbl. (iii) In the 2.5Mb region of Chr15, there are three TADs in the WT (top panel in Fig. 5e). The first and the third TADs have an epigenetic switch at only one boundary (blue dashed line), and the expectation is that they would not be preserved upon Nipbl removal. However, the boundary probabilities show that the TADs have physical boundaries in both, and thus they are preserved. Taken together, the results in Fig. 5 show that by combining the epigenetic switches (Hi-C data is sufficient) and the boundary probabilities (3D structures are required), one can account for a number of P-TADs.

Single-cell structural change upon cohesin depletion

Finally, we asked whether the HIPPS method captures the 3D structural changes in cohesin-depleted cell at the single-cell level. To this end, we compare the structures obtained using HIPPS with the imaging data [40], which examined the consequences of ΔRAD21 in HCT-116. We used HIPPS, with Hi-C contact map as input on the same genomic region as in the experiment [40], to generate the 3D structures. The results of our calculations for a 2.5Mbp in Chr21 (34.6-37.1Mb) region from HCT-116 cell line for the WT and ΔRAD21 are presented in Fig. III. The distance maps were calculated from the calculated 3D structures using the HIPPS method (Fig. IIIa). The mean distance maps for the WT and ΔRAD21 are shown on the left and right panels in Fig. IIIb. Similar results for Chr 4 in mouse liver cell (Chr2 in HCT-116) are displayed in Appendix Fig. 9 (Appendix Fig. 10).

Several conclusions may be drawn from the results in Fig. III. (i) There are large variations in the distance matrices and the domain boundary locations/strengths from cell to cell (Fig. IIIc and d). This finding is in excellent agreement with imaging data [40]. (ii) In both experiments and our calculations, there are TAD-like structures at the single level even after RAD21 is removed (see the right panel in Fig. IIIc for the theoretical predictions). TAD-like structures in single cells with and without cohesin have also been found using the strings and binders polymer model [12]. (iii) The calculated boundary strength distribution (blue histogram in the left panel in Fig. IIId) for the WT is in reasonable agreement with the measured distribution (purple histogram from [40]). Similarly, the calculated and measured boundary strength distributions for ΔRAD21 cells are also in good agreement (right panel in Fig. IIId). Just as in experiments [40], we find that the distributions of boundary strength are the same in the WT and in cells without RAD21. (iv) We also find that the theoretically calculated average locus-dependent boundary probability is in very good agreement with the reported experimental data (compare the curves in the left panel in Fig. IIIe for the WT and the ones on the right for 6.RAD21 cells).

P-TADs are due to enhancer/promoter interactions

Cohesin is thought to directly or indirectly regulate enhancer-promoter (E-P) interactions. However, a recent Micro-C experiment discovered that E-P and promoter-promoter (P-P) interactions are, to a large extent, insensitive to acute depletion of cohesin [48]. It has been previously shown that E-P/P-P interactions form one or multiple self-associating domains, strips that extend from domain borders and loop-like structures at their intersections at a finer scale [49–51]. Inspired by the recent finding [48], we explored if P-TADs that arise in the absence of epigenetic switches, are required for the maintenance of finer-scale E-P and P-P interactions. We analyzed the Micro-C data [48] in order to shed light on this issue. The left panel in Fig. 7a shows cohesin-associated (green dashed line) and cohesin-independent (blue dashed line) TAD structures (defined using Topdom) in the WT cells. In the latter case, the E-P and P-P loops (blue circles) are at the boundary of the TADs even in the absence of no epigenetic switch, implying that it is a domain that is needed for E-P or P-P communication. Interestingly, the TADs were also conserved upon cohesin loss (right panel in Fig. 7a). Analyses of the 3D structures (Fig. 7b) reveal that the TADs with E-P/P-P loops have strong physical boundaries sans cohesin. Fig. 7c shows an example of a TAD with both E-P/P-P loops and cohesin/CTCF loops at the boundary in the WT cells that is retained after cohesin deletion, and is associated with prominent boundary peaks. We propose that only a subset of TADs are conserved, potentially for functional reasons.

The statistical analyses of all the P-TADs observed in the Micro-C contact maps across all the chromosomes show that 525 out of 1536 P-TADs have E-P/P-P loops that coincide with their boundaries (Fig. 8). Taken together, our observations suggest that maintenance of E-P/P-P interactions could be the origin of the P-TADs even if there are no epigenetic switches. It is worth emphasizing that these conclusions can only be obtained by analyzing the 3D structures, which we calculated from the Micro-C contact maps using the HIPPS method [45] that does not rely on polymer simulations.

Discussion

By analyzing the experimental Hi-C data [26, 34], we first showed that upon cohesin loss a non-negligible fraction of TADs is preserved, which was not previously noticed. To examine the factors that control the P-TADs, we then performed polymer simulations of two chromosomes in the presence and absence of cohesin-associated CTCF loops from the GM12878 cell line. The polymer simulation results were used to generate hypotheses for the emergence of P-TADs, which were used to explain the major findings reported in the experiments in mouse liver and HCT-116 cells [26, 34]. The simulations showed, switches in the epigenetic states across the TAD boundary account for a large fraction of P-TADs. Even in the absence of epigenetic switches, P-TADs could be preserved, as revealed by the presence of physical boundaries in the 3D structures.

Rather than perform a number of time-consuming polymer simulations, we used the data-driven approach [45] that generates three-dimensional chromosome structures rapidly and accurately using the Hi-C contact maps as input. Analyses of the calculated structures, with and without Nipbl or RAD21, showed that A-type loci form larger spatial clusters after cohesin removal, consistent with enhancement of compartments inferred from Hi-C contact maps (Appendix Fig. 4). Most of the P-TADs, with epigenetic switches in the contact maps, have prominent peaks in the boundary probabilities in both WT and cohesin-depleted cells. An important conclusion from this striking finding is that not only micro-phase separation on the larger scale, LC but also some special TADs on a shorter scale, LTAD are encoded in the epigenetic sequence.

Remarkably, the conclusion that there are cell-to-cell variations in the distance maps, noted in imaging experiments [40], are affirmed in the calculated 3D structures. This finding is significant because (1) only a limited number of loci can be directly imaged whereas Hi-C data can be routinely generated at higher resolution, (2) the number of Hi-C data on various cell types and species currently is far greater than that obtained from imaging data.

Let us summarize the novel results, which sets our work apart from previous insightful studies [26, 34, 35]. (1) We showed by analyzing the Hi-C data for mouse liver and HCT-116 cell lines that a non-negligible fraction of TADs is preserved, which set in motion our detailed investigations. (2) Then, using polymer simulations on a different cell type (GM12878), we generated quantitative insights (epigenetic switches as well as structural basis) for the preservation of TADs. Although not emphasized, we showed that deletion of cohesin in the GM12878 cell line also leads to P-TADs, a prediction that suggests that P-TADs may be “universal”. (3) Rather than perform time consuming polymer simulations, we calculated 3D structures directly from Hi-C data for the mouse liver and HCT-116 cell lines, which provided a structural basis for TAD preservation. (4) The 3D structures also showed how TAD-like features appear at the single-cell level, which is in accord with imaging experiments [40]. (5) Finally, we suggest that P-TADs may be linked to the maintenance of enhancerpromoter and promoter-promoter interactions by calculating the 3D structures using the recent Micro-C data [48].

Comments on the Methods

In order to explore the factors that control the P-TADs, there are two assumptions. (1) The results of the Hi-C experiments are taken at face value in the sense. We view this as an assumption because errors in the Hi-C readouts may be difficult to evaluate even though such experiments are invaluable [52]. (2) The TADs were identified using TopDom, one of many TAD callers. A recent survey [53] shows that, although the finding that TADs are hierarchically organized domains is robust, there are substantial variations in the identification of these domains predicted by different methods. Although TopDom fairs reasonably well in comparison to other methods, there is no guarantee that it identifies the TAD location or the number of TADs accurately. It is only for convenience that we used TopDom as the reference to which the results using the boundary probabilities are compared. (3) Because the prediction of 3D structures using the HIPPS method does not require extensive polymer simulations, it can be used to predict the structural changes for chromosomes that are subject to large-scale perturbations. The excellent agreement between the HIPPS calculations and imaging experiments further bolsters the power of our approach.